---
title: "pairs_timeseries"
format: html
editor_options: 
  chunk_output_type: console
---

source("functions.R")

```{r}
library(lubridate)
library(tidyverse)
library(terra)
library(dplyr)
source("D:/2025_Osa_Conservation/GIS_projects/OSA_LST_analyses_R/02_lst_analyses/functions.R")
```

# Raw LST data cleaning and visualization

```{r}
# Read into a tiblle
path <- "D:/2025_Osa_Conservation/GIS_projects/OSA_LST_analyses_R/00_data/data_selection/"

csv <- read_csv(paste0(path, "pairs_timeseries_results.csv"))


```

```{r}
# at high pairs one pair (poly 30) has switched landcover values
# test <- high_csv |>
#   mutate(landcover = case_when(poly_id == 30 & landcover == 30 ~ 10, poly_id == 30 & landcover == 10 ~ 30, TRUE ~ landcover))
# 
# # write.csv(test, "data/high_pairs_timeseries_results.csv",row.names = FALSE)
```

```{r}
# Clean the raw data

df <- clean_lst_data(csv)
df

```

### Plot raw

```{r}
# Plot the timeseries
p <- plot_raw (df, "0-1800")
p

# Save the plots
# Map(ggsave,
  #   filename = c("media/p1.png", "media/p2.png", "media/p3.png"),
  #   plot = list(p1, p2, p3),
  #   width = 250,
  #   height = 150,
  #   units = "mm"
  # )
```

### Facet grids for each separate polygon

```{r}

plot_facet<- function (df, elev_range) {
  ggplot(df, aes(x = date, y = lst, color = landcover)
  ) +
    # geom_point(alpha = 0.4) +            # optional: show the raw points
    geom_jitter(size = 0.5, alpha = 0.25) +
    geom_smooth(method = "loess",        # "loess" for local smoothing
                span = 0.1,              # adjust smoothness (0-1)
                se = FALSE,
                fill = "#4e4f52",
                linewidth = 0.6) +               # display confidence interval
    scale_colour_manual(
      values = c("darkgreen", "sienna2")
    ) +
    scale_y_continuous(limits = c(5, 35), breaks = seq(5, 35, by = 5)
    ) +
    guides(x = guide_axis(angle = 45)) +
    # scale_x_date(date_breaks = "2 months",  date_labels = "%Y %b")+
    # scale_y_continuous(limits = c(5, 40), breaks = seq(5, 40, by = 5))+
    annotate("rect", xmin = as.Date("2023-11-01"), xmax = as.Date("2024-04-01"), ymin = -Inf, ymax = Inf,         alpha = .2
    ) +
    # annotate("text", x = as.Date("2023-12-01"), y = 30, label = "El Niño"
    # ) +
    labs(
      x = "Date", 
      y = "LST (°C)", 
      title = paste("Daily LST time-series for forest and open land at",elev_range, "m")
    ) + facet_wrap(~poly_id, ncol = 5) +
    theme_minimal()
    
}



# facet <- plot_facet(df, "0-1800")
# ggsave(
#   "facet.png",
#     facet,
#     width = 250,
#     height = 900,
#     units = "mm"
# )
```

# Monthly LST 

Calculate monthly LST for all points

```{r}
# function to get teh monthly value for each point
monthly_lst <- function(df) {
  df |>
    mutate(
      year  = year(date),
      month = month(date)
    ) |>
    group_by(landcover,year, month) |>
    summarise(
      lst_monthly_mean = mean(lst, na.rm = TRUE),
      n_points = n(),    # optional (how many records per month)
    ) |>
    ungroup() |>
    # optional: create a proper first-of-month date
    mutate(month_date = as.Date(sprintf("%04d-%02d-01", year, month)))
}

monthly <- monthly_lst(df)
monthly

hottest_months <- df_monthly |>
  filter(landcover == "open_land") |>
  group_by(year) |>
  slice_max(lst_monthly_mean, n = 1, with_ties = FALSE) |>
  ungroup()

hottest_months
# # results
# landcover  year month lst_monthly_mean n_points month_date
# open_land  2021     4             19.5     1216 2021-04-01
# open_land  2022     3             20.1     1280 2022-03-01
# open_land  2023     4             21.0     1226 2023-04-01
# open_land  2024     4             22.3     1216 2024-04-01
    
```

Plot monthly values

```{r}
library(scales)
source("D:/2025_Osa_Conservation/GIS_projects/OSA_LST_analyses_R/02_lst_analyses/functions.R")

p_month <- plot_monthly(monthly,"0-1800m")
p_month
```

## Change analyses with the bfastmonitor() function

Looking at Oct 2020 - Oct 2022 as a reference (no el Nino conditions) and detecting whether any breaks occur after that

```{r}
library(bfast)

df_monthly <- monthly

# Choose one landcover - forest or open land
open_land_monthly_summarize <- df_monthly |>  filter(landcover == "open_land")
open_land_monthly_summarize

forest_monthly_summarize <- df_monthly |>  filter(landcover == "forest")
forest_monthly_summarize
# ------------------------------------------------------------------
# stl --> seasonal trend decomposition with loess
ts_o <- ts(open_land_monthly_summarize$lst_monthly_mean, frequency = 12, start = c(2020,10))
plot(stl(ts_o,"per"))

ts_f <- ts(forest_monthly_summarize$lst_monthly_mean, frequency = 12, start = c(2020,10))
plot(stl(ts_f,"per"))

# ------------------------------------------------------------------
# bfast

plot(bfast(ts_o, season = "harmonic", breaks = 1))
plot(bfast(ts_f, season = "harmonic", breaks = 1))

# ------------------------------------------------------------------
# bfastmonitor

bfm <- function(df){
  ts <- ts(df$lst_monthly_mean, frequency = 12, start = c(2020,10))
  bfm <- bfastmonitor(ts, formula = response ~ harmon, start = c(2022,10))
}

plot(bfm(open_land_monthly_summarize),ylab = "LST", ylim = c(14,28), sub = "Open Land at 0-1800m")
(bfm(open_land_monthly_summarize))$magnitude

plot(bfm(forest_monthly_summarize),ylab = "LST", ylim = c(14,28), sub = "Forest at 0-1800m")
(bfm(forest_monthly_summarize))$magnitude
# # |>
# #   select(c(lst))
#   
# ts <- ts(open_land_monthly_summarize$lst, frequency = 12, start = c(2020,10))
# ts
# 
# 
# bfm <- bfastmonitor(ts, formula = response ~ harmon, start = c(2022,10))
# plot(bfm,ylab = "LST", ylim = c(16,28), sub = "Open Land at 0-100m")
# 
# ts(1:10, frequency = 4, start = c(1959,1), end = 1962) # 2nd Quarter of 1959
```

```{r}
# 
# source("functions.R")
# 
# Map(plot_bfast_change,
#     list(df_monthly, df_monthly),
#     c("forest", "open_land"),
#     c("0-1800m", "0-1800m")
#     )

```

# Monthly Anomaly 

## Mean across all points

```{r}
# df <- low_df

plot_monthly_anom <- function (df, df_monthly, elev_range) {

  #--------------------------------------
  # Get the monthly mean for the first two reference years for each pair points
  monthly_ref_means <- df |>
  filter(date >= as.Date("2020-10-01"),
         date <= as.Date("2022-10-31")
  ) |>
  # extract month number
  mutate(month = month(date)
  ) |>
  # compute the mean for each month (across all years in the window)
  group_by(landcover, month, poly_id) |>
  summarise(
    ref_mean = mean(lst, na.rm = TRUE),
    .groups = "drop"
  )
  
  #--------------------------------------
  #Get the monthly mean for each pair accross the whole period
  df_monthly <- df |>
    mutate(
      year  = year(date),
      month = month(date)
    ) |>
    group_by(landcover,year, month, poly_id) |>
    summarise(
      lst_monthly_mean = mean(lst, na.rm = TRUE),
      n_points = n(),    # optional (how many records per month)
    ) |>
    ungroup() |>
    # optional: create a proper first-of-month date
    mutate(month_date = as.Date(sprintf("%04d-%02d-01", year, month)))

  #--------------------------------------
  df_monthly_anom <- df_monthly |>
  left_join(monthly_ref_means, by = c("landcover", "month", "poly_id")) |>
  mutate(anomaly = lst_monthly_mean - ref_mean )
  
#--------------------------------------
  plot_anom<- function (df, elev_range) {
  ggplot(
    df,
    aes(
      x = month_date,
      y = anomaly,
      color = landcover)
    )+
    # geom_line(
    #   aes(color = landcover),
    #   linewidth = 0.6
    # ) +
    geom_jitter(size = 0.6, alpha = 0.5) +
    geom_smooth(method = "loess",# "loess" for local smoothing
                span = 0.1,      # adjust smoothness (0-1)
                se = FALSE,
                fill = "#4e4f52",
                linewidth = 1) +
    scale_colour_manual(
      values = c("darkgreen", "sienna2")
    ) +
    scale_colour_manual(
      values = c("darkgreen", "sienna2")
    ) +
    scale_y_continuous(
      limits = c(-3, 8), 
      breaks = seq(-2, 8, by = 2)
      )+
    annotate("rect", xmin = as.Date("2023-11-01"), xmax = as.Date("2024-04-01"), ymin = -Inf, ymax = Inf, alpha = .2
    ) +
    annotate("text", x = as.Date("2023-12-01"), y = 28, label = "El Niño"
    ) +
    geom_hline(
      yintercept = 0, 
      color = "#4e4f52", 
      linetype = "dashed", 
      linewidth = 0.8) +
      annotate("rect", xmin = as.Date("2023-11-01"), xmax = as.Date("2024-04-01"), ymin = -Inf, ymax = Inf, alpha = .2
    )+
    labs(
      x = "Date",
      y = "LST anomaly (°C)",
      title = paste("Monthly LST anomaly of forest and open land at",elev_range, "m")
    )+
    theme_minimal()
}

plot_anom(df_monthly_anom,elev_range)

}


plot_monthly_anom (df, df_monthly, "0-1800m")
```

## Facets: Pair wise

```{r}
# df <- low_df

plot_monthly_anom <- function (df, df_monthly, elev_range) {

  #--------------------------------------
  # Get the monthly mean for the first two reference years for each pair points
  monthly_ref_means <- df |>
  filter(date >= as.Date("2020-10-01"),
         date <= as.Date("2022-10-31")
  ) |>
  # extract month number
  mutate(month = month(date)
  ) |>
  # compute the mean for each month (across all years in the window)
  group_by(landcover, month, poly_id) |>
  summarise(
    ref_mean = mean(lst, na.rm = TRUE),
    .groups = "drop"
  )
  
  #--------------------------------------
  #Get the monthly mean for each pair accross the whole period
  df_monthly <- df |>
    mutate(
      year  = year(date),
      month = month(date)
    ) |>
    group_by(landcover,year, month, poly_id) |>
    summarise(
      lst_monthly_mean = mean(lst, na.rm = TRUE),
      n_points = n(),    # optional (how many records per month)
    ) |>
    ungroup() |>
    # optional: create a proper first-of-month date
    mutate(month_date = as.Date(sprintf("%04d-%02d-01", year, month)))

  #--------------------------------------
  df_monthly_anom <- df_monthly |>
  left_join(monthly_ref_means, by = c("landcover", "month", "poly_id")) |>
  mutate(anomaly = lst_monthly_mean - ref_mean )
  
#--------------------------------------
  plot_anom<- function (df, elev_range) {
  ggplot(
    df,
    aes(
      x = month_date,
      y = anomaly,
      color = landcover)
    )+
    geom_line(
      aes(color = landcover),
      linewidth = 0.6
    ) +
    geom_jitter(size = 0.6, alpha = 0.25) +
    # geom_smooth(method = "loess",        # "loess" for local smoothing
    #             span = 0.5,              # adjust smoothness (0-1)
    #             se = FALSE,
    #             fill = "#4e4f52",
    #             linewidth = 0.9) +               # display confidence interval
    scale_colour_manual(
      values = c("darkgreen", "sienna2")
    ) +
    scale_colour_manual(
      values = c("darkgreen", "sienna2")
    ) +
    scale_y_continuous(
      limits = c(-3, 8), 
      breaks = seq(-2, 8, by = 2)
      )+
    annotate("rect", xmin = as.Date("2023-11-01"), xmax = as.Date("2024-04-01"), ymin = -Inf, ymax = Inf, alpha = .2
    ) +
    annotate("text", x = as.Date("2023-12-01"), y = 28, label = "El Niño"
    ) +
    geom_hline(
      yintercept = 0, 
      color = "#4e4f52", 
      linetype = "dashed", 
      linewidth = 0.8) +
      annotate("rect", xmin = as.Date("2023-11-01"), xmax = as.Date("2024-04-01"), ymin = -Inf, ymax = Inf, alpha = .2
    )+
    labs(
      x = "Date",
      y = "LST anomaly (°C)",
      title = paste("Monthly LST anomaly of forest and open land at",elev_range, "m")
    ) + facet_wrap(~poly_id, ncol = 5) +
    theme_minimal()
}

plot_anom(df_monthly_anom,elev_range)

}


# Map(plot_monthly_anom,
#     list(low_df, mid_df, high_df),
#     list(low_monthly, mid_monthly, high_monthly),
#     c("0-100m","600-700", "1300-1500")
# )

facet_anom <- plot_monthly_anom (df, df_monthly, "0-1800m")


# ggsave(
#   "facet_anom.png",
#     facet_anom,
#     width = 250,
#     height = 900,
#     units = "mm"
# )
```

# Cooling calculation

Calculate the cooling for each pait (with identical poly_id) thorughout the study period. Cooling is expressed as the LST difference between open land and forest.

```{r}
library(readr)

df_cool <- df |>
  pivot_wider(
    id_cols = c(poly_id, date),
    names_from = landcover,
    values_from = lst
  ) |>
  mutate(
    forest_cooling = open_land - forest
  )

df_cool
# Save the cooling table for fast reuse
write_csv(df_cool, "df_cool.csv")

```

## Compare cooling for hottest months

```{r}
# function to calculate the monthly cooling
monthly_cooling <- function(df) {
  df |>
    mutate(
      year  = year(date),
      month = month(date)
    ) |>
    group_by(poly_id, year, month) |>
    summarise(
      cooling_monthly_mean = mean(forest_cooling, na.rm = TRUE),
      n_points = n(),    # optional (how many records per month)
    ) |>
    ungroup() |>
    # optional: create a proper first-of-month date
    mutate(month_date = as.Date(sprintf("%04d-%02d-01", year, month)))
}

# creat a df for monthlu cooling
df_cool_month <- monthly_cooling(df_cool)

# filter to include only the hottest months (based on monthly lst for open land)
df_cool_month <- df_cool_month |>
  filter(
    (year == 2021 & month == 4) |
    (year == 2022 & month == 3) |
    (year == 2023 & month == 4) |
    (year == 2024 & month == 4)
  )

df_cool_month
```

### Plot the data using boxplots

```{r}
# Libraries
library(tidyverse)
library(hrbrthemes)
library(viridis)

df_cool_month %>%
  ggplot( aes(x = factor(year), y=cooling_monthly_mean, fill=year)) +
    geom_boxplot() +
    scale_fill_viridis( alpha=0.6) +
  scale_y_continuous(breaks = seq(floor(min(df_cool_month$cooling_monthly_mean, na.rm = TRUE)),
                                ceiling(max(df_cool_month$cooling_monthly_mean, na.rm = TRUE)),
                                by = 2)) +
    geom_jitter(color="black", size=0.4, alpha=0.9) +
    theme_ipsum() +
    theme(
      legend.position="none",
      plot.title = element_text(size=11)
    ) +
  xlab("")+
  ylab("Forest cooling (°C)")
```

### Plot cooling time-series: Mean across all pairs

```{r}

plot_forest_cooling <- function(df, elev_range) {
  
df<-df_cool


ggplot(df_cool, aes(x = date, y = forest_cooling)
    )+
    geom_jitter(size = 0.5, alpha = 0.25
    )+
    geom_smooth(method = "loess",        # "loess" for local smoothing
                span = 0.1,              # adjust smoothness (0-1)
                se = FALSE,
                fill = "#4e4f52",
                linewidth = 0.8,
                color = 'black'
    )+
   geom_smooth(method=lm, color = 'darkred'
    )+
    geom_hline(
      yintercept = 0, 
      color = "#4e4f52", 
      linetype = "dashed", 
      linewidth = 0.8
    )+
    scale_y_continuous(
      limits = c(-2.5, 10), 
      breaks = seq(-2.5, 10, by = 2.5)
      )+
    annotate("rect", xmin = as.Date("2023-11-01"), xmax = as.Date("2024-04-01"), ymin = -Inf, ymax = Inf, alpha = .2
    )+
    labs(
      x = "Date",
      y = "Forest Cooling (°C)",
      title = paste("Forest induced cooling at",elev_range, "m")
    )+
  theme_minimal()
}

plot_forest_cooling(df,"0-1800m")
    
```

### Plot cooling: For each pair

```{r}

plot_forest_cooling_pairs <- function(df, elev_range){ 
  df<-df_cool

  ggplot(df_cool, aes(x = date, y = forest_cooling)
    )+
    geom_jitter(size = 0.5, alpha = 0.25
    )+
    geom_smooth(method = "loess",        # "loess" for local smoothing
                  span = 0.1,              # adjust smoothness (0-1)
                  se = FALSE,
                  fill = "#4e4f52",
                  linewidth = 0.8,
                color = 'black'
    )+
    geom_smooth(method=lm, color = 'darkred'
    )+
    geom_hline(
      yintercept = 0, 
      color = "#4e4f52", 
      linetype = "dashed", 
      linewidth = 0.8) +
      annotate("rect", xmin = as.Date("2023-11-01"), xmax = as.Date("2024-04-01"), ymin = -Inf, ymax = Inf, alpha = .2
    )+
    scale_y_continuous(
      limits = c(-2.5, 10), 
      breaks = seq(-2.5, 10, by = 2.5)
    )+
    labs(
      x = "Date",
      y = "Forest Cooling (°C)",
      title = paste("Forest induced cooling at",elev_range, "m")
    )+
    facet_wrap(~poly_id, ncol = 5
    )+
    theme_bw()
  }

facet_cool <- plot_forest_cooling_pairs(df,"0-1800m")

ggsave(
  "facet_cool.png",
    facet_cool,
    width = 250,
    height = 900,
    units = "mm"
)
```

## Compare cooling accross elevations

Add elevation for each poly_id based on the mean elevation value from forest and open land

```{r}

# load the points data
pts <- vect("D:/2025_Osa_Conservation/GIS_projects/OSA_LST_analyses_R/00_data/data_selection/pairs_32617.shp")
# convert it to a dataframe
df_pts <- as.data.frame(pts)

# set the landcover as a factor, as now it is just a code (10 and 30)
df_pts$landcover <- factor(
  df_pts$landcover,
  levels = c(10, 30),
  labels = c("forest", "open_land")
)

df_pts

# calculate the mean elevation for each poly_id
poly_elv <- df_pts |>
  pivot_wider(
    id_cols = poly_id,
    names_from = landcover,
    values_from = elv
   )|>
  mutate(
    poly_elv = (open_land + forest)/2
  )

poly_elv
```

Join the table containing the elevation for each polygon with the table containing the cooling values

```{r}
df_cool_month <- df_cool_month %>%
  left_join(
    poly_elv %>% select(poly_id, poly_elv),
    by = "poly_id"
  )
df_cool_month
```

### Plot the data

```{r}
p4 <- ggplot(df_cool_month, aes(x=poly_elv, y=cooling_monthly_mean)) +

  geom_point() +

  geom_smooth(method=lm ,formula = y ~ poly(x, 2), color="red", fill="#69b3a2", se=TRUE) +

  scale_y_continuous(breaks = seq(floor(min(df_cool_month$cooling_monthly_mean, na.rm = TRUE)),

                                ceiling(max(df_cool_month$cooling_monthly_mean, na.rm = TRUE)),

                                by = 2))+

  facet_wrap(~year, ncol = 4) +

  theme_ipsum() 

p4
```

```{r}
# linear trend + confidence interval
p3 <- ggplot(df_cool_month, aes(x=poly_elv, y=cooling_monthly_mean)) +
  geom_point() +
  geom_smooth(method=lm , color="red", fill="#69b3a2", se=TRUE) +
  scale_y_continuous(breaks = seq(floor(min(df_cool_month$cooling_monthly_mean, na.rm = TRUE)),
                                ceiling(max(df_cool_month$cooling_monthly_mean, na.rm = TRUE)),
                                by = 2))+
    labs(
      x = "Elevation (m)",
      y = "Forest Cooling (°C)",
      title = "Linear model") +
  facet_wrap(~year, ncol = 4) +
  theme_ipsum() 

p3
# Non linear - polynomial
p4 <- ggplot(df_cool_month, aes(x=poly_elv, y=cooling_monthly_mean)) +
  geom_point() +
  geom_smooth(method=lm ,formula = y ~ poly(x, 2), color="red", fill="#69b3a2", se=TRUE) +
  scale_y_continuous(breaks = seq(floor(min(df_cool_month$cooling_monthly_mean, na.rm = TRUE)),
                                ceiling(max(df_cool_month$cooling_monthly_mean, na.rm = TRUE)),
                                by = 2))+
  labs(
      x = "Elevation (m)",
      y = "Forest Cooling (°C)",
      title = "Polynomial model") +
  facet_wrap(~year, ncol = 4) +
  theme_ipsum() 

p4

# Non linear - loess
p4 <- ggplot(df_cool_month, aes(x=poly_elv, y=cooling_monthly_mean)) +
  geom_point() +
  geom_smooth(method=loess, color="red", fill="#69b3a2", se=TRUE) +
  scale_y_continuous(breaks = seq(floor(min(df_cool_month$cooling_monthly_mean, na.rm = TRUE)),
                                ceiling(max(df_cool_month$cooling_monthly_mean, na.rm = TRUE)),
                                by = 2))+
    labs(
      x = "Elevation (m)",
      y = "Forest Cooling (°C)",
      title = "Loess model") +
  facet_wrap(~year, ncol = 4) +
  theme_ipsum() 

p4
```

### Statistical test

```{r}
# First: reshape your data into paired format for 2021 and 2024 comparison
df_wide <- df_cool_month %>%
  filter(year %in% c(2021, 2022, 2023, 2024)) %>%
  select(poly_id, year, cooling_monthly_mean) %>%
  pivot_wider(names_from = year, values_from = cooling_monthly_mean)

df_wide

# Quick sanity check
df_wide <- df_wide %>%
  mutate(diff = `2024` - `2021`)

summary(df_wide$diff)
hist(df_wide$diff)

# !!!! difference distribution seems slightly skewed. Check what to do in this case

```

#### Hypothesis (important)

Because you *expect 2024 \> 2021*, this is a **one-sided paired test**.

#### Null hypothesis (H₀):

H0:μ2024−2021=0H_0: \mu_{2024 - 2021} = 0H0​:μ2024−2021​=0

(no change in LST)

#### Alternative hypothesis (H₁):

H1:μ2024−2021>0H_1: \mu_{2024 - 2021} > 0H1​:μ2024−2021​\>0

(2024 is warmer on average)

```{r}
# Run the paired t-test in R
t.test(df_wide$`2024`, df_wide$`2021`,
       paired = TRUE,
       alternative = "greater")

# Run the paired t-test in R
t.test(df_wide$`2024`, df_wide$`2022`,
       paired = TRUE,
       alternative = "greater")

# Run the paired t-test in R
t.test(df_wide$`2024`, df_wide$`2023`,
       paired = TRUE,
       alternative = "greater")
```
