---
title: "pairs_timeseries"
format: html
---

source("functions.R")

```{r}
library(lubridate)
library(tidyverse)
source("functions.R")
```

```{r}
# Read into a tiblle
path <- "D:/2025_Osa_Conservation/GIS_projects/OSA_LST_analyses_R/00_data/"
low_csv <- read_csv(paste0(path, "low_pairs_timeseries_results.csv"))
mid_csv <- read_csv(paste0(path,"mid_pairs_timeseries_results.csv"))
high_csv <- read_csv(paste0(path,"high_pairs_timeseries_results.csv"))

```

```{r}
# at high pairs one pair (poly 30) has switched landcover values
# test <- high_csv |>
#   mutate(landcover = case_when(poly_id == 30 & landcover == 30 ~ 10, poly_id == 30 & landcover == 10 ~ 30, TRUE ~ landcover))
# 
# # write.csv(test, "data/high_pairs_timeseries_results.csv",row.names = FALSE)
```

```{r}
# Clean the raw data

low_df <- clean_lst_data(low_csv)
low_df
mid_df <- clean_lst_data(mid_csv)
high_df <- clean_lst_data(high_csv)

```

```{r}
# Plot the timeseries
p1 <- plot_raw (low_df, "0-100")
p2 <- plot_raw (mid_df, "600-700")
p3 <- plot_raw (high_df, "1300-1500")

p1
p2
p3

# Save the plots
# Map(ggsave,
  #   filename = c("media/p1.png", "media/p2.png", "media/p3.png"),
  #   plot = list(p1, p2, p3),
  #   width = 250,
  #   height = 150,
  #   units = "mm"
  # )
```

### Facet grids for each separate polygon

```{r}

plot_facet<- function (df, elev_range) {
  ggplot(df, aes(x = date, y = lst, color = landcover)
  ) +
    # geom_point(alpha = 0.4) +            # optional: show the raw points
    geom_jitter(size = 0.5, alpha = 0.25) +
    geom_smooth(method = "loess",        # "loess" for local smoothing
                span = 0.1,              # adjust smoothness (0-1)
                se = FALSE,
                fill = "#4e4f52",
                linewidth = 0.6) +               # display confidence interval
    scale_colour_manual(
      values = c("darkgreen", "sienna2")
    ) +
    scale_y_continuous(limits = c(5, 35), breaks = seq(5, 35, by = 5)
    ) +
    guides(x = guide_axis(angle = 45)) +
    # scale_x_date(date_breaks = "2 months",  date_labels = "%Y %b")+
    # scale_y_continuous(limits = c(5, 40), breaks = seq(5, 40, by = 5))+
    annotate("rect", xmin = as.Date("2023-11-01"), xmax = as.Date("2024-04-01"), ymin = -Inf, ymax = Inf,         alpha = .2
    ) +
    # annotate("text", x = as.Date("2023-12-01"), y = 30, label = "El Niño"
    # ) +
    labs(
      x = "Date", 
      y = "LST (°C)", 
      title = paste("Daily LST time-series for forest and open land at",elev_range, "m")
    ) + facet_wrap(~poly_id, ncol = 5) +
    theme_minimal()
    
}

# p <- plot_facet(low_df, "0-100")
# p
Map(plot_facet,
    list(low_df, mid_df, high_df),
    c("0-100", "600-700", "1300-1500")) 

plot_facet(high_df, "1300-1500")
```

## Monthly LST plot

```{r}

low_monthly <- monthly(low_df)
mid_monthly <- monthly(mid_df)
high_monthly <- monthly(high_df)

Map(plot_monthly,
    list(low_monthly, mid_monthly, high_monthly),
    c("0-100m", "600-700m", "1300-1500m")
    )
```

## Change analyses with the bfastmonitor() function

Looking at Oct 2020 - Oct 2022 as a reference (no el Nino conditions) and detecting whether any breaks occur after that

```{r}
library(bfast)

df_monthly <- low_monthly

# Choose one landcover - forest or open land
open_land_monthly_summarize <- df_monthly |>  filter(landcover == "open_land")
open_land_monthly_summarize

forest_monthly_summarize <- df_monthly |>  filter(landcover == "forest")
forest_monthly_summarize
# ------------------------------------------------------------------
# stl --> seasonal trend decomposition with loess
ts_o <- ts(open_land_monthly_summarize$lst_monthly_mean, frequency = 12, start = c(2020,10))
plot(stl(ts_o,"per"))

ts_f <- ts(forest_monthly_summarize$lst_monthly_mean, frequency = 12, start = c(2020,10))
plot(stl(ts_f,"per"))

# ------------------------------------------------------------------
# bfast

plot(bfast(ts_o, season = "harmonic", breaks = 1))
plot(bfast(ts_f, season = "harmonic", breaks = 1))

# ------------------------------------------------------------------
# bfastmonitor

bfm <- function(df){
  ts <- ts(df$lst_monthly_mean, frequency = 12, start = c(2020,10))
  bfm <- bfastmonitor(ts, formula = response ~ harmon, start = c(2022,10))
}

plot(bfm(open_land_monthly_summarize),ylab = "LST", ylim = c(14,28), sub = "Open Land at 0-100m")
(bfm(open_land_monthly_summarize))$magnitude

plot(bfm(forest_monthly_summarize),ylab = "LST", ylim = c(14,28), sub = "Forest at 0-100m")
(bfm(forest_monthly_summarize))$magnitude
# # |>
# #   select(c(lst))
#   
# ts <- ts(open_land_monthly_summarize$lst, frequency = 12, start = c(2020,10))
# ts
# 
# 
# bfm <- bfastmonitor(ts, formula = response ~ harmon, start = c(2022,10))
# plot(bfm,ylab = "LST", ylim = c(16,28), sub = "Open Land at 0-100m")
# 
# ts(1:10, frequency = 4, start = c(1959,1), end = 1962) # 2nd Quarter of 1959
```

```{r}
# BFAST Example

plot(harvest, ylab = "NDVI")

(rdist <- 10/length(harvest))
# ratio of distance between breaks (time steps) and length of the time series

fit <- bfast(harvest, h = rdist, season = "harmonic", max.iter = 2, breaks = 1)
plot(fit)

plot(fit, ANOVA = TRUE)
## plot the trend component and identify the break with
## the largest magnitude of change
plot(fit, type = "trend", largest = TRUE)

## plot all the different available plots
plot(fit, type = "all")

```

```{r}

source("functions.R")

Map(plot_bfast_change,
    list(low_monthly, low_monthly),
    c("forest", "open_land"),
    c("0-100m", "0-100m")
    )

Map(plot_bfast_change,
    list(mid_monthly, mid_monthly),
    c("forest", "open_land"),
    c("600-700m", "600-700m"),
    list(c(10,24), c(10,24))
    )

Map(plot_bfast_change,
    list(high_monthly, high_monthly),
    c("forest", "open_land"),
    c("1300-1500m", "1300-1500m"),
    list(c(6,20),c(6,20))
    )
```

## Monthly Anomaly: Mean across all points

```{r}
# df <- low_df

plot_monthly_anom <- function (df, df_monthly, elev_range) {

  #--------------------------------------
  # Get the monthly mean for the first two reference years for each pair points
  monthly_ref_means <- df |>
  filter(date >= as.Date("2020-10-01"),
         date <= as.Date("2022-10-31")
  ) |>
  # extract month number
  mutate(month = month(date)
  ) |>
  # compute the mean for each month (across all years in the window)
  group_by(landcover, month, poly_id) |>
  summarise(
    ref_mean = mean(lst, na.rm = TRUE),
    .groups = "drop"
  )
  
  #--------------------------------------
  #Get the monthly mean for each pair accross the whole period
  df_monthly <- df |>
    mutate(
      year  = year(date),
      month = month(date)
    ) |>
    group_by(landcover,year, month, poly_id) |>
    summarise(
      lst_monthly_mean = mean(lst, na.rm = TRUE),
      n_points = n(),    # optional (how many records per month)
    ) |>
    ungroup() |>
    # optional: create a proper first-of-month date
    mutate(month_date = as.Date(sprintf("%04d-%02d-01", year, month)))

  #--------------------------------------
  df_monthly_anom <- df_monthly |>
  left_join(monthly_ref_means, by = c("landcover", "month", "poly_id")) |>
  mutate(anomaly = lst_monthly_mean - ref_mean )
  
#--------------------------------------
  plot_anom<- function (df, elev_range) {
  ggplot(
    df,
    aes(
      x = month_date,
      y = anomaly,
      color = landcover)
    )+
    # geom_line(
    #   aes(color = landcover),
    #   linewidth = 0.6
    # ) +
    geom_jitter(size = 0.6, alpha = 0.5) +
    geom_smooth(method = "loess",# "loess" for local smoothing
                span = 0.1,      # adjust smoothness (0-1)
                se = FALSE,
                fill = "#4e4f52",
                linewidth = 0.6) +
    scale_colour_manual(
      values = c("darkgreen", "sienna2")
    ) +
    scale_colour_manual(
      values = c("darkgreen", "sienna2")
    ) +
    scale_y_continuous(
      limits = c(-3, 8), 
      breaks = seq(-2, 8, by = 2)
      )+
    annotate("rect", xmin = as.Date("2023-11-01"), xmax = as.Date("2024-04-01"), ymin = -Inf, ymax = Inf, alpha = .2
    ) +
    annotate("text", x = as.Date("2023-12-01"), y = 28, label = "El Niño"
    ) +
    geom_hline(
      yintercept = 0, 
      color = "#4e4f52", 
      linetype = "dashed", 
      linewidth = 0.8) +
      annotate("rect", xmin = as.Date("2023-11-01"), xmax = as.Date("2024-04-01"), ymin = -Inf, ymax = Inf, alpha = .2
    )+
    labs(
      x = "Date",
      y = "LST anomaly (°C)",
      title = paste("Monthly LST anomaly of forest and open land at",elev_range, "m")
    )+
    theme_minimal()
}

plot_anom(df_monthly_anom,elev_range)

}


Map(plot_monthly_anom,
    list(low_df, mid_df, high_df),
    list(low_monthly, mid_monthly, high_monthly),
    c("0-100m","600-700", "1300-1500")
)
```

# Monthly anomaly: Pair wise

```{r}
# df <- low_df

plot_monthly_anom <- function (df, df_monthly, elev_range) {

  #--------------------------------------
  # Get the monthly mean for the first two reference years for each pair points
  monthly_ref_means <- df |>
  filter(date >= as.Date("2020-10-01"),
         date <= as.Date("2022-10-31")
  ) |>
  # extract month number
  mutate(month = month(date)
  ) |>
  # compute the mean for each month (across all years in the window)
  group_by(landcover, month, poly_id) |>
  summarise(
    ref_mean = mean(lst, na.rm = TRUE),
    .groups = "drop"
  )
  
  #--------------------------------------
  #Get the monthly mean for each pair accross the whole period
  df_monthly <- df |>
    mutate(
      year  = year(date),
      month = month(date)
    ) |>
    group_by(landcover,year, month, poly_id) |>
    summarise(
      lst_monthly_mean = mean(lst, na.rm = TRUE),
      n_points = n(),    # optional (how many records per month)
    ) |>
    ungroup() |>
    # optional: create a proper first-of-month date
    mutate(month_date = as.Date(sprintf("%04d-%02d-01", year, month)))

  #--------------------------------------
  df_monthly_anom <- df_monthly |>
  left_join(monthly_ref_means, by = c("landcover", "month", "poly_id")) |>
  mutate(anomaly = lst_monthly_mean - ref_mean )
  
#--------------------------------------
  plot_anom<- function (df, elev_range) {
  ggplot(
    df,
    aes(
      x = month_date,
      y = anomaly,
      color = landcover)
    )+
    geom_line(
      aes(color = landcover),
      linewidth = 0.6
    ) +
    geom_jitter(size = 0.6, alpha = 0.25) +
    # geom_smooth(method = "loess",        # "loess" for local smoothing
    #             span = 0.5,              # adjust smoothness (0-1)
    #             se = FALSE,
    #             fill = "#4e4f52",
    #             linewidth = 0.9) +               # display confidence interval
    scale_colour_manual(
      values = c("darkgreen", "sienna2")
    ) +
    scale_colour_manual(
      values = c("darkgreen", "sienna2")
    ) +
    scale_y_continuous(
      limits = c(-3, 8), 
      breaks = seq(-2, 8, by = 2)
      )+
    annotate("rect", xmin = as.Date("2023-11-01"), xmax = as.Date("2024-04-01"), ymin = -Inf, ymax = Inf, alpha = .2
    ) +
    annotate("text", x = as.Date("2023-12-01"), y = 28, label = "El Niño"
    ) +
    geom_hline(
      yintercept = 0, 
      color = "#4e4f52", 
      linetype = "dashed", 
      linewidth = 0.8) +
      annotate("rect", xmin = as.Date("2023-11-01"), xmax = as.Date("2024-04-01"), ymin = -Inf, ymax = Inf, alpha = .2
    )+
    labs(
      x = "Date",
      y = "LST anomaly (°C)",
      title = paste("Monthly LST anomaly of forest and open land at",elev_range, "m")
    ) + facet_wrap(~poly_id, ncol = 5) +
    theme_minimal()
}

plot_anom(df_monthly_anom,elev_range)

}


Map(plot_monthly_anom,
    list(low_df, mid_df, high_df),
    list(low_monthly, mid_monthly, high_monthly),
    c("0-100m","600-700", "1300-1500")
)
```

# Forest cooling calculation: Mean across all pairs

```{r}

plot_forest_cooling <- function(df, elev_range) {
df_cool <- df |>
  pivot_wider(
    id_cols = c(poly_id, date),
    names_from = landcover,
    values_from = lst
  ) |>
  mutate(
    forest_cooling = open_land - forest
  )


ggplot(df_cool, aes(x = date, y = forest_cooling)
    )+
    geom_jitter(size = 0.5, alpha = 0.25
    )+
    geom_smooth(method = "loess",        # "loess" for local smoothing
                span = 0.1,              # adjust smoothness (0-1)
                se = FALSE,
                fill = "#4e4f52",
                linewidth = 0.8,
                color = 'black'
    )+
   geom_smooth(method=lm, color = 'darkred'
    )+
    geom_hline(
      yintercept = 0, 
      color = "#4e4f52", 
      linetype = "dashed", 
      linewidth = 0.8
    )+
    scale_y_continuous(
      limits = c(-2.5, 10), 
      breaks = seq(-2.5, 10, by = 2.5)
      )+
    annotate("rect", xmin = as.Date("2023-11-01"), xmax = as.Date("2024-04-01"), ymin = -Inf, ymax = Inf, alpha = .2
    )+
    labs(
      x = "Date",
      y = "Forest Cooling (°C)",
      title = paste("Forest induced cooling at",elev_range, "m")
    )+
  theme_minimal()
}

Map(plot_forest_cooling,
    list(low_df, mid_df, high_df),
    c("0-100m","600-700", "1300-1500")
    )
```

# Forest cooling calculation: For each pair

```{r}
plot_forest_cooling_pairs <- function(df, elev_range){ 
  df_cool <- df |>
  pivot_wider(
    id_cols = c(poly_id, date),
    names_from = landcover,
    values_from = lst
  ) |>
  mutate(
    forest_cooling = open_land - forest
  )


  ggplot(df_cool, aes(x = date, y = forest_cooling)
    )+
    geom_jitter(size = 0.5, alpha = 0.25
    )+
    geom_smooth(method = "loess",        # "loess" for local smoothing
                  span = 0.1,              # adjust smoothness (0-1)
                  se = FALSE,
                  fill = "#4e4f52",
                  linewidth = 0.8,
                color = 'black'
    )+
    geom_smooth(method=lm, color = 'darkred'
    )+
    geom_hline(
      yintercept = 0, 
      color = "#4e4f52", 
      linetype = "dashed", 
      linewidth = 0.8) +
      annotate("rect", xmin = as.Date("2023-11-01"), xmax = as.Date("2024-04-01"), ymin = -Inf, ymax = Inf, alpha = .2
    )+
    scale_y_continuous(
      limits = c(-2.5, 10), 
      breaks = seq(-2.5, 10, by = 2.5)
    )+
    labs(
      x = "Date",
      y = "Forest Cooling (°C)",
      title = paste("Forest induced cooling at",elev_range, "m")
    )+
    facet_wrap(~poly_id, ncol = 5
    )+
    theme_bw()
  }

Map(plot_forest_cooling_pairs,
    list(low_df, mid_df, high_df),
    c("0-100m","600-700", "1300-1500")
    )
```
