---
title: "lst_elevation_low_veg_vs_forest"
format: html
---

# Analyse LST difference between pairs of nearby low vegetation and forest pixels at different elevation gradients

## 1) Prepare the data

Data needed:

-   landover raster

-   elevation raster

```{r}
library(terra)
library(ggplot2)
library(dplyr)
```

```{r}

# load landcover data (categorical raster)
ldc <- rast("D:/2025_Osa_Conservation/DATA_YN/data_major_landocvers_LST_comparison/landcover_210_pure_YN_5367_major_landcov.tif")
# load elevation data (continuous raster)
elv <- rast("D:/2025_Osa_Conservation/DATA_YN/data_major_landocvers_LST_comparison/elevation_210_bilinear_5367_major_landcov.tif") 

print(ldc)
print(elv)
```

```{r}
plot(elv)
```

```{r}

# filter only classes 10 and 30 in the landcover layer (forest and low vegetation)
# Alternatively, faster way:
keep_vals <- c(10, 30)
ldc_filtered <- mask(ldc, ldc %in% keep_vals, maskvalues=FALSE)

plot(ldc_filtered, col = c("forestgreen","goldenrod"))
```

## 2) Find the areas that contain forest and low vegetation pairs

Get a 4x4 pixels rolling window to find:

-   **A)** max (30 landcover)

-   **B)** min (10 landocover)

```{r}
# aggregate by factor of 4 (210 m → 840 m)
A <- aggregate(ldc_filtered, fact = 4, fun = max, na.rm = TRUE)
B <- aggregate(ldc_filtered, fact = 4, fun = min, na.rm = TRUE)
# C <- aggregate(elv, fact = 5, fun = max, na.rm = TRUE)
# D <- aggregate(elv, fact = 5, fun = min, na.rm = TRUE)
```

```{r}
plot(A,col = c("forestgreen","goldenrod"))
plot(B,col = c("forestgreen","goldenrod"))
```

## Raster calculator

**If A + B = 40 then 1, else 0**

This intends to find all the agregated cells (4x4) where both land covers are found (10 from min and 30 from max). This will give us the areas where we can make comparison

```{r}
# # create mask: if A + B = 40 and D - C > 100 then 1 else 0
# mask <- ((A + B == 40) & (C - D < 200)) * 1

# create mask: if A + B = 40 and D - C > 100 then 1 else 0
mask <- ((A + B == 40)) * 1

plot(mask)

#writeRaster(mask, "mask_forest_low_veg_pairs.tif", overwrite = TRUE)
```

Convert the mask pixels (pixels containing pairs) to polygons with unique ids

```{r}
#Convert each pixel to a polygon
mask1 <- mask
mask1[mask1 != 1] <- NA   # all other pixels become NA

polygons <- as.polygons(mask1, dissolve = FALSE)
# Optionally, assign an ID to each polygon
polygons$ID <- 1:nrow(polygons)
# Inspect
polygons
```

Save them locally (optional)

```{r}
# Save as GeoJSON
writeVector(polygons, "mask_forest_low_veg_pair.geojson", filetype = "GeoJSON", overwrite = TRUE)
```

## 3) Extract paired points

To do this we will first convert all the landcover pixels which fall in the masked raster into points

```{r}
#first, convert the masked raster to the same resolution as the original landcover raster
mask <- resample(mask, ldc_filtered, method = "near")  # nearest neighbor

# then apply the mask to the landcover data. This will leave only the areas where we have pairs.
ldc_mask <- ldc_filtered*mask
plot(ldc_mask, col = c("white","forestgreen","goldenrod"))

# turn all zeros to NULL
ldc_mask[ldc_mask == 0] <- NA

#convert all remaining pixels to points
ldc_pts <- as.points(ldc_mask, na.rm = TRUE)

#write the id of each polygon to the points
ldc_pts$poly_id <- extract(polygons, ldc_pts)[, "ID"]

#Attach landcover class from the raster
ldc_pts$landcover <- extract(ldc_mask, ldc_pts)[,2]  # 2nd column = raster values

print(ldc_pts)


```

Add attributes of elevation and UID

```{r}

ldc_pts$elv <- extract(elv, ldc_pts)[,2]
ldc_pts$uid <- seq_len(nrow(ldc_pts)) 
ldc_pts
nrow(ldc_pts)
```

Now we will filter only those elevation categories for which we intend to do our LST analyses

```{r}

# Extract attribute table as data frame
df <- as.data.frame(ldc_pts)

# Leave only those points at the chosen elevation category
df_filtered <- df %>%
  filter( 
    (elv >= 0 & elv <= 100) |
    (elv >= 600 & elv <= 700) |
    (elv >= 1300 & elv <= 1500)
  )
  
# Now subset the original vector to keep only those rows
pts_filtered <- ldc_pts[df_filtered$uid, ]   # or use ID column if present
pts_filtered
```

Now we will remove the duplicate points within one polygon.

```{r}
# Get only the attribute table (does not touch geometry)
att <- values(pts_filtered)

# Find duplicates by poly_id + landcover
keep <- !duplicated(att[c("poly_id", "landcover")])

# Subset the SpatVector
pts_unique <- pts_filtered[keep, ]
pts_unique

```

And remove any single points, where there is not a pair anymore

```{r}
# Get only the attribute table (does not touch geometry)
att_u <- values(pts_unique)

# Count how many points per poly_id
poly_counts <- table(att_u$poly_id)

# Keep only poly_ids with more than one point
keep_ids <- names(poly_counts[poly_counts > 1]) 

# Subset the SpatVector
pts_unique <- pts_unique[att_u$poly_id %in% keep_ids, ]

pts_unique
```

## 4) Get 10 pairs of spatially balanced points for each elevation

With the code above we have obtained:

-   218 pairs at low elevation (0-100m)

-   28 pairs at medium elevation (600-700m)

-   20 pairs at high elevation (1300-1500)

Now we want to extract 10 pairs of spatially balanced samples for each elevation class

**4.1) Divide the data in our three categories**

We will extract the polygons we defined in step 2, which contain our pairs. Then we will select 10 polygons based on the spatially balanced method (grts) for each elev category. After that we will select only those points from points_unique that have the same poly_id as our spatially balanced polygons.

```{r}

# Extract attributes
att <- values(pts_unique)

# Devide the points data in their elevation category. 
pts_low  <- pts_unique[att$elv <= 100, ]
pts_mid  <- pts_unique[att$elv >= 600 & att$elv <= 700, ]
pts_high <- pts_unique[att$elv >= 1300, ]

print(nrow(pts_low)/2)
print(nrow(pts_mid)/2)
print(nrow(pts_high)/2)

# Function to extract polygons matching point poly_ids
get_polys_for_points <- function(pts, polys, pts_id = "poly_id", poly_id = "ID") {
  # Extract unique IDs from points
  ids <- unique(values(pts)[[pts_id]])
  # Subset polygons using their ID field
  polys_subset <- polys[values(polys)[[poly_id]] %in% ids, ]
  return(polys_subset)
}

# Get the polygons which have those points
# In addition we take the centers as points and convert them from spat vector to sf object. This extra steps make it easier when working with the grts function in spsurvey later
polys_low_sf <- st_as_sf(centroids(get_polys_for_points(pts_low, polygons)))
polys_mid_sf <- st_as_sf(centroids(get_polys_for_points(pts_mid, polygons)))
polys_high_sf <- st_as_sf(centroids(get_polys_for_points(pts_high, polygons)))

print(nrow(polys_low_sf))
print(nrow(polys_mid_sf))
print(nrow(polys_high_sf))


```

**4.2) Get spatially balanced samples**

```{r}
library(spsurvey)
set.seed(51)
```

```{r}
# Now we will extract the spatially balanced data with the grts function from spsurvey for each elevation class

# # TEST
# poly_high_grts <- (grts(polys_high_sf, n_base = 10))
# plot(poly_high_grts, polys_high_sf, key.width = lcm(3))

# We add $sites_base to convert the result in sf object
# Apply to all the point datasets
poly_low_grts <- (grts(polys_low_sf, n_base = 10))$sites_base
poly_mid_grts <- (grts(polys_mid_sf, n_base = 10))$sites_base
poly_high_grts <- (grts(polys_high_sf, n_base = 10))$sites_base



```

Extract the ids of the spatially balanced polygons

```{r}
ids_low_grts <- poly_low_grts$ID
ids_mid_grts <- poly_mid_grts$ID
ids_high_grts <- poly_high_grts$ID

# print(length(unique(ids_low_grts)))
# print(length(unique(ids_mid_grts)))
# print(length(unique(ids_high_grts)))
```

```{r}
pts_low_grts <- pts_low[values(pts_low)$poly_id %in%ids_low_grts, ]
pts_mid_grts <- pts_mid[values(pts_mid)$poly_id %in%ids_mid_grts, ]
pts_high_grts <- pts_high[values(pts_high)$poly_id %in%ids_high_grts, ]

```

## 5) Export the data

```{r}

# # all pairs generated
# writeVector(pts_unique, "DATA/pairs_timseries_v260123_1432.shp",
#             filetype = "ESRI Shapefile", overwrite = TRUE)

# spatially balanced 10 pairs for each elevation category
writeVector(pts_low_grts, "DATA/pairs_timseries_low_10.shp",
            filetype = "ESRI Shapefile", overwrite = TRUE)

writeVector(pts_mid_grts, "DATA/pairs_timseries_mid_10.shp",
            filetype = "ESRI Shapefile", overwrite = TRUE)

writeVector(pts_high_grts, "DATA/pairs_timseries_high_10.shp",
            filetype = "ESRI Shapefile", overwrite = TRUE)
```
