---
title: "lst_elevation_low_veg_vs_forest"
format: html
---

# Analyse LST difference between pairs of nearby low vegetation and forest pixels at different elevation gradients

## 1) Prepare the data

Data needed:

-   landover raster

-   elevation raster

```{r}
library(terra)
library(ggplot2)
library(dplyr)
```

```{r}

# load landcover data (categorical raster)
ldc <- rast("D:/Conservation_Biology/2025_Osa_Conservation/DATA_YN/data_major_landocvers_LST_comparison/landcover_210_pure_YN_5367_major_landcov.tif")
# load elevation data (continuous raster)
elv <- rast("D:/Conservation_Biology/2025_Osa_Conservation/DATA_YN/data_major_landocvers_LST_comparison/elevation_210_bilinear_5367_major_landcov.tif") 

print(ldc)
print(elv)
```

```{r}
plot(elv)
```

```{r}

# filter only classes 10 and 30 in the landcover layer (forest and low vegetation)
# Alternatively, faster way:
keep_vals <- c(10, 30)
ldc_filtered <- mask(ldc, ldc %in% keep_vals, maskvalues=FALSE)

plot(ldc_filtered, col = c("forestgreen","goldenrod"))
```

## 2) Find the areas that contain forest and low vegetation pairs

Get a 3x3 pixels rolling window to find:

-   **A)** max (30 landcover)

-   **B)** min (10 landocover)

-   **C)** max (elevation)

-   **D)** min (elevation)

```{r}
# aggregate by factor of 3 (10 m → 30 m)
A <- aggregate(ldc_filtered, fact = 5, fun = max, na.rm = TRUE)
B <- aggregate(ldc_filtered, fact = 5, fun = min, na.rm = TRUE)
# C <- aggregate(elv, fact = 5, fun = max, na.rm = TRUE)
# D <- aggregate(elv, fact = 5, fun = min, na.rm = TRUE)
```

```{r}
plot(A,col = c("forestgreen","goldenrod"))
plot(B,col = c("forestgreen","goldenrod"))
```

## Raster calculator

**If A + B = 40 and C - D \< 100 then 1, else 0**

This intends to find all the cells where both land covers are found (10 from min and 30 from max) and where the difference in elevation is not greater than 100m. This will give us the areas where we can make comparison

```{r}
# # create mask: if A + B = 40 and D - C > 100 then 1 else 0
# mask <- ((A + B == 40) & (C - D < 200)) * 1

# create mask: if A + B = 40 and D - C > 100 then 1 else 0
mask <- ((A + B == 40)) * 1

plot(mask)

#writeRaster(mask, "mask_forest_low_veg_pairs.tif", overwrite = TRUE)
```

Convert the mask pixels (pixels containing pairs) to polygons with unique ids

```{r}
#Convert each pixel to a polygon
mask1 <- mask
mask1[mask1 != 1] <- NA   # all other pixels become NA

polygons <- as.polygons(mask1, dissolve = FALSE)
# Optionally, assign an ID to each polygon
polygons$ID <- 1:nrow(polygons)
# Inspect
polygons
```

Save them locally (optional)

```{r}
# Save as GeoJSON
writeVector(polygons, "mask_forest_low_veg_pair.geojson", filetype = "GeoJSON", overwrite = TRUE)
```

## 3) Extract paired points

To do this we will first convert all the landcover pixels which fall in the masked raster into points

```{r}
#first, convert the masked raster to the same resolution as the original landcover raster
mask <- resample(mask, ldc_filtered, method = "near")  # nearest neighbor

# then apply the mask to the landcover data. This will leave only the areas where we have pairs.
ldc_mask <- ldc_filtered*mask
plot(ldc_mask, col = c("white","forestgreen","goldenrod"))

# turn all zeros to NULL
ldc_mask[ldc_mask == 0] <- NA

#convert all remaining pixels to points
ldc_pts <- as.points(ldc_mask, na.rm = TRUE)

#write the id of each polygon to the points
ldc_pts$poly_id <- extract(polygons, ldc_pts)[, "ID"]

# for each landcover class, leave only one point with the same id
#Attach landcover class from the raster
ldc_pts$landcover <- extract(ldc_mask, ldc_pts)[,2]  # 2nd column = raster values

print(ldc_pts)


```

Now we will remove the duplicate points within one polygon.

```{r}
# Get only the attribute table (does not touch geometry)
att <- values(ldc_pts)

# Find duplicates by poly_id + landcover
keep <- !duplicated(att[c("poly_id", "landcover")])

# Subset the SpatVector
pts_unique <- ldc_pts[keep, ]

```

```{r}
pts_unique$elv <- extract(elv, pts_unique)[,2]
pts_unique$uid <- seq_len(nrow(pts_unique))
pts_unique
```

Keep only those points within 100m

```{r}

# Extract attribute table as data frame
df <- as.data.frame(pts_unique)

# Compute elevation range within each poly_id (pairs)
df_filtered <- df %>%
  group_by(poly_id) %>%
  mutate(elv_diff = max(elv) - min(elv)) %>%
  filter(elv_diff <= 100,
    all(elv >= 0 & elv <= 100)||(elv >= 600 & elv <= 700)||(elv >= 1600 & elv <= 1700)) %>%
  ungroup()

# Now subset the original vector to keep only those rows
pts_filtered <- pts_unique[df_filtered$uid, ]   # or use ID column if present
pts_filtered
```

```{r}

# # Save as shapefile
writeVector(pts_unique, "pairs_timseries.shp",
            filetype = "Shapefile", overwrite = TRUE)
```

```{r}
pts_unique
```

## 4) Check their distribution accorss elevations

\# Create df from attribute table (without geometry)

df \<- as.data.frame(pts_unique, geom = FALSE)

df

```{r}
# Extract raster values
vals <- extract(elv, pts_unique)

# attach elevation to attributes
pts_unique$elev <- vals[, 2]

pts_unique

```

```{r}
df <- as.data.frame(pts_unique)
```

```{r}

# If landcover is categorical, convert to factor
df$landcover <- factor(df$landcover,
                       levels = c(10, 30),
                       labels = c("Forest","Low vegetation", "Urban")) 

# Plot distribution of elevation across landcovers
ggplot(df, aes(x = landcover, y = elev, fill = landcover)) +
  geom_boxplot() +
  labs(x = "Landcover type", y = "Elevation (m)", 
       title = "Elevation distribution across landcover classes") +
  theme_minimal()
```

```{r}
library(mapview)
library(raster)

ldc_mask <- raster(ldc_mask)
# Define categories and colors
vals <- c(0,10,30)
cols = c("white","forestgreen","goldenrod")

mapview(ldc_mask, col.regions = cols, at = vals)
```
