---
title: "Untitled"
format: html
---

## 1) Prepare the data

Data needed:

-   landover raster

-   elevation raster

```{r}
library(terra)
library(ggplot2)
library(dplyr)
library(viridis)
library(sf)
library(spsurvey)
```

```{r}

# load landcover data (categorical raster)
ldc <- rast("D:/Conservation_Biology/2025_Osa_Conservation/DATA_YN/data_major_landocvers_LST_comparison/landcover_210_pure_YN_5367_major_landcov.tif")
# load elevation data (continuous raster)
elv <- rast("D:/Conservation_Biology/2025_Osa_Conservation/DATA_YN/data_major_landocvers_LST_comparison/elevation_210_bilinear_5367_major_landcov.tif") 

print(ldc)
print(elv)
```

```{r}

# filter only classes 10 and 30 in the landcover layer (forest and low vegetation)
# Alternatively, faster way:

ldc <- mask(ldc, ldc %in% c(10, 30), maskvalues=FALSE)

print(unique(ldc))
plot(ldc, col = c("forestgreen","goldenrod"))
```

# Get spatially balanced dataset that is representative accross all elevations

-   First create your elevation-landcover strata

-   The convert each pixel center to a point

-   Then apply the grts function

We start by classifying the elevation in strata

```{r}
elv_classes <- classify(elv, rcl = matrix(c(
    -4, 300, 1,
    300, 600, 2,
    600, 900, 3,
    900, 1200, 4,
    1200, 1500, 5,
    1500, 1800, 6,
    1800, 3354, 7
), ncol = 3, byrow = TRUE))

print(unique(elv_classes))
```

Then we combine it with the landcover raster in such a way that we get unique strata for each landcover at each elevation

```{r}
stack_raster <- ldc * 10 + elv_classes
# Get unique values
# Print them
print(unique(stack_raster))

plot(stack_raster, type = "classes", col = viridis(14, option = "plasma"))
```

Next step is to convert all pixel centers to points:

```{r}
#convert all remaining pixels to points
ldc_pts <- as.points(stack_raster, na.rm = TRUE)
ldc_pts
```

```{r}
# Count how many rows there are in a ceratian starta e.g. Band_1 == 307
print(sum(values(ldc_pts)$Band_1 == 107, na.rm = TRUE))
```

Our points are now in"SpatVector" But in order to use it with the grts function in spsurvey we need to convert it to sf object.

```{r}
pts_sf <- st_as_sf(ldc_pts)

# If not already factor/character, convert:
pts_sf$stratum <- as.factor(pts_sf$Band_1)

```

Now we can extract the spatially balanced points with the grts function

```{r}
set.seed(51)
```

```{r}
strata_n = c('101' = 100,'102' = 100, '103' = 100, '104' = 100, '105' = 100, '106' = 100, '301' = 100,'302' = 100, '303' = 100, '304' = 100, '305' = 100, '306' = 100) #, '107' = 100, '307' = 36)

strat_eqprob = grts(pts_sf, n_base = strata_n, stratum_var = "Band_1", mindis = 210)

strat_eqprob
```

```{r}
str(strat_eqprob)
```

```{r}
plot(strat_eqprob, pts_sf, key.width = lcm(3))
```

```{r}
# convert to sf object and then to spatvector
strat_eqprob_sf <-strat_eqprob$sites_base
```

```{r}

names(strat_eqprob_sf)
```

```{r}
strat_eqprob_sf <- strat_eqprob_sf[, !(names(strat_eqprob_sf) %in% c("siteuse", "replsite", "wgt", "ip", "caty", "sframe_stratum", "lat_WGS84", "lon_WGS84" ))]

```

```{r}
st_crs(strat_eqprob_sf) -> n5367
```

```{r}
strat_eqprob_sf$landcover <- substr(strat_eqprob_sf$stratum, 1, 2)
```

```{r}
st_write(strat_eqprob_sf, "data/points_low_veg_forest_5367_v2.gpkg",append=FALSE)
```
