---
?---
title: "lst_elevation_low_veg_vs_forest"
format: html
---

# Analyse LST difference between pairs of nearby low vegetation and forest pixels at different elevation gradients

## 1) Prepare the data

Data needed:

-   landover raster

-   elevation raster

```{r}
library(terra)
library(ggplot2)
library(dplyr)
```

# Prepossessing of input data

## CRS projectoins

Reproject the data to the same projection. For Costa Rica, we have chosen EPSG: 32EPSG:32617EPSG:32617EPSG:3261761

```{r}

# load landcover data (categorical raster)
ldc <- rast("D:/2023_OSA_Conservation_LULC_internship/DATA_final_results - zenodo/8_Amistosa/8_Amostosa_landcover_YN.tif")
ldc <- project(ldc, "EPSG:32617")
writeRaster(
  ldc, 
  "D:/2025_Osa_Conservation/GIS_projects/OSA_LST_analyses_R/00_data/data_selection/landcover_32617.tif", 
  overwrite=TRUE)

# load elevation data (continuous raster)

elv <- rast("D:/2025_Osa_Conservation/DATA_YN/elevation_30_YN.tif") 
elv <- project(elv, "EPSG:32617")
writeRaster(
  elv, 
  "D:/2025_Osa_Conservation/GIS_projects/OSA_LST_analyses_R/00_data/data_selection/elevation_30_32617.tif", 
  overwrite=TRUE)

# water extent
mwe <- rast("D:/2025_Osa_Conservation/GIS_projects/OSA_LST_analyses_R/00_data/data_selection/extent_90W_10Nv1_4_2021.tif")

mwe <- project(mwe, "EPSG:32617")
# We will crop the mwe dataset, to save memory, since it is too big
mwe_crop <- crop(x = mwe, y = aoi)
writeRaster(
  mwe_crop, 
  "D:/2025_Osa_Conservation/GIS_projects/OSA_LST_analyses_R/00_data/data_selection/water_extent_30_32617.tif", 
  overwrite=TRUE)
# mwe <- crop(x = mwe, y = aoi)
# crs(ldc)
# crs(elv)
# crs(ch)
# crs(mwe)

# Load open street map rivers
rivers <- vect("D:/2025_Osa_Conservation/GIS_projects/OSA_LST_analyses_R/00_data/hotosm_cri_waterways_lines_shp/hotosm_cri_waterways_lines_shp.shp")
rivers <- project(rivers, "EPSG:32617")

# crop to aoi extent
rivers_crop <- crop(x = rivers, y = aoi)
rivers_crop 
crs(rivers_crop)

# save the projected vector
writeVector(
  rivers_crop,
  "D:/2025_Osa_Conservation/GIS_projects/OSA_LST_analyses_R/00_data/data_selection/rivers_osm_32617.shp"
)
```

## 

```{r}
# area of interest (aoi)
aoi <- vect("D:/2025_Osa_Conservation/DATA/Amistosa_climate_hub_32617.shp")

ldc <- rast("D:/2025_Osa_Conservation/GIS_projects/OSA_LST_analyses_R/00_data/data_selection/landcover_10_2023_32617.tif")

elv <- rast("D:/2025_Osa_Conservation/GIS_projects/OSA_LST_analyses_R/00_data/data_selection/elevation_30_32617.tif")

# load carbon height (Planet product 30m)
ch <- rast("D:/2025_Osa_Conservation/GIS_projects/OSA_LST_analyses_R/00_data/data_selection/canopy_height_30_2023_32617.tif") 
ch <- crop(x = ch, y = aoi)

# maximum water extent
mwe <- rast("D:/2025_Osa_Conservation/GIS_projects/OSA_LST_analyses_R/00_data/data_selection/water_extent_30_32617.tif")

# open street map rivers
rivers <- vect("D:/2025_Osa_Conservation/GIS_projects/OSA_LST_analyses_R/00_data/data_selection/rivers_osm_32617.shp")
```

```{r}
plot(elv, col = (terra.pal=terrain.colors(10)), range=c(1,3000))
plot(ldc)
plot(ch,range=c(0,30))
plot(mwe,col=c("light gray", "blue"))
plot(rivers,col="blue")
plot(aoi, add = TRUE)
```

```{r}

res(ldc)
res(elv)
res(ch)
res(mwe)
```

## Reclassify canopy height to forest and non-forest

```{r}
# Reclassify values into categories --> everything from 0-3 m is open land, and everything above 15 meters is forest
rc <- matrix(c(
  -0.5, 2, 30,
  14.9, 40, 10
), ncol=3, byrow=TRUE)

ch_cat <- classify(ch, rc, others=NA)
plot(ch_cat,col = c("forestgreen","goldenrod"))
```

## Create a mask from landcover and water

Remove the following classes:

-   palm oil

-   pineapple

-   urban

-   mangroves

-   water

```{r}
# filter only classes 10 and 30 in the landcover layer (forest and low vegetation)
# Alternatively, faster way:
ldc_mask <- ifel(ldc %in% c(10, 30, 110), 1, NA)
water_mask <- ifel(mwe>0,NA,1)

# Resample the mask layers to have the same resolution as the canopy heigh layer.
ldc_mask_30 <- resample(ldc_mask, ch, min)
water_mask_30 <- resample(water_mask, ch, min)

# plot(water_mask, col = "dark grey")
plot(water_mask_30, col = "dark grey")

plot(ldc_mask_30, col = "dark grey")
plot(water_mask_30*ldc_mask_30, col = "dark grey")
```

Apply the mask to the forest non-forest raster

```{r}
ldc_pairs <- ch_cat*water_mask_30*ldc_mask_30
plot(ldc_pairs, col = c("forestgreen","goldenrod"))

# writeRaster(
#  ldc_pairs, 
#   "D:/2025_Osa_Conservation/GIS_projects/OSA_LST_analyses_R/00_data/data_selection/ldc_pairs.tif", 
#   overwrite=TRUE)
```

## Option 1: Aggregate cells to 210m and check for pure pixels

Here we check whether at least 90% of the 30m pixels are of a single class.

```{r}
# function to calculate the mode and how often the mode has occured
mode_with_count <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(c(NA, NA))

  tab <- table(x)
  m <- tab[which.max(tab)]

  c(mode = as.numeric(names(m)), count = as.numeric(m))
}

# aggregate to 210m pixel resolution --> two bands, one with the mode, one wiht the majority count
r_agg <- aggregate(
  ldc_pairs,
  fact = 7,
  fun = mode_with_count
)

plot(r_agg)
# ---------------------------------------
# show the areas where there have been at least 19 cells (out of 21) with the mode
r19 <- ifel(r_agg[[2]] >= 47, 1, NA)

plot(r19)
# ------------------------------
# make sure not a single water pixel falls in the aggregated cells
water_mask_210 <- resample(water_mask_30, r19, min)
r19<- r19*water_mask_210
plot(r19)
# ------------------------------
ldc_pure <- r19*r_agg[[1]]
plot(ldc_pure, col = c("forestgreen","goldenrod"))
# plot(ldc_210, col = c("grey", "forestgreen","goldenrod"))
writeRaster(
 ldc_pure,
  "D:/2025_Osa_Conservation/GIS_projects/OSA_LST_analyses_R/00_data/data_selection/ldc_pure5.tif",
  overwrite=TRUE)
```

Another approach to check whether 100% of 30m pixels in a 210m pixel are a single class, by checking if the minimum and maximum from the aggregate fucntion are the same.

```{r}
ldc_min <- aggregate(
  ldc_pairs,
  fact = 7,
  fun = min
)

ldc_max <- aggregate(
  ldc_pairs,
  fact = 7,
  fun = max
)

plot(ldc_pairs,col = c("forestgreen","goldenrod"))
plot(ldc_min,col = c("forestgreen","goldenrod"))
plot(ldc_max,col = c("forestgreen","goldenrod"))

ldc_pure <- ifel((ldc_max-ldc_min)==0, ldc_max, NA)
plot(ldc_pure,col = c("forestgreen","goldenrod"))

writeRaster(
 ldc_pure,
  "D:/2025_Osa_Conservation/GIS_projects/OSA_LST_analyses_R/00_data/data_selection/ldc_pure4.tif",
  overwrite=TRUE)
```

## Option 2: Use rolling window

```{r}
# The matrix representing the search window
w <- matrix(1, 7, 7)

# function to calculate the mode and how often the mode has occured
mode_with_count <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(c(NA, NA))

  tab <- table(x)
  m <- tab[which.max(tab)]

  c(mode = as.numeric(names(m)), count = as.numeric(m))
}

pure_check <- focal(ldc_pairs, w = w, fun = mode_with_count)

plot(pure_check[[2]])
```

```{r}
# Check if mode occurence is above 48 (97%)
r97 <- ifel(pure_check[[2]] >= 48, 1, NA)

plot(r97)
# ------------------------------
# make sure not a single water pixel falls in the aggregated cells
r97<- r97*water_mask_30
plot(r97)
# ------------------------------
ldc_pure <- r97*pure_check[[1]]
plot(ldc_pure, col = c("forestgreen","goldenrod"))
# plot(ldc_210, col = c("grey", "forestgreen","goldenrod"))
writeRaster(
 ldc_pure,
  "D:/2025_Osa_Conservation/GIS_projects/OSA_LST_analyses_R/00_data/data_selection/ldc_pure6.tif",
  overwrite=TRUE)
```

## Add elevation variables (elevation and slope)

Data should be considered unreliable over sloped terrain that is over 20 degrees. 

```{r}

# First resample elevatioon to 30 m ldc_pure layer
elv_30 <- crop(elv, ext(ldc_pure))
elv_30 <- resample(elv, ldc_pure, mean)


# Calculate the slope
slope <- terrain(elv_30, v = "slope", unit = "degrees")


# The matrix representing the search window
w <- matrix(1, 7, 7)

# # Calculte only the mode occurences
# mode_count <- function(x) {
#   x <- x[!is.na(x)]
#   if(length(x) == 0) return(NA)
# 
#   tab <- table(x)
#   max(tab)
# }
# function to calculate the mode and how often the mode has occured
slope_mean <- focal(slope, w = w, fun = mean)



# mask out areas over 20 degrees of slope
slope_check <- ifel(slope_mean >= 20, NA, 1)

plot(slope)
plot(slope_mean)
plot(slope_check, col = "dark grey")


# Apply the slope check to the ldc_pure raster
ldc_slope <- ldc_pure*slope_check
plot(ldc_slope, col = c("forestgreen","goldenrod"))
# 
writeRaster(
 ldc_slope,
  "D:/2025_Osa_Conservation/GIS_projects/OSA_LST_analyses_R/00_data/data_selection/ldc_pure7.tif",
  overwrite=TRUE)

# crs(elv_30)
# crs(ldc_pure)
# 
# ext(elv_30)
# ext(ldc_pure)
# 
# res(elv_30)
# res(ldc_pure)
# 
# # Check coordinate reference system (CRS)
# crs(ldc_pure)
# crs(slope_check)
# 
# # Check extent
# ext(ldc_pure)
# ext(slope_check)
# 
# # Check resolution
# res(ldc_pure)
# res(slope_check)

```

## 2) Find the areas that contain forest and low vegetation pairs

Get a 4x4 pixels rolling window to find:

-   **A)** max (30 landcover)

-   **B)** min (10 landocover)

```{r}
# aggregate by factor of 4 (210 m → 840 m)
A <- aggregate(ldc_filtered, fact = 4, fun = max, na.rm = TRUE)
B <- aggregate(ldc_filtered, fact = 4, fun = min, na.rm = TRUE)
# C <- aggregate(elv, fact = 5, fun = max, na.rm = TRUE)
# D <- aggregate(elv, fact = 5, fun = min, na.rm = TRUE)
```

```{r}
plot(A,col = c("forestgreen","goldenrod"))
plot(B,col = c("forestgreen","goldenrod"))
```

## Raster calculator

**If A + B = 40 then 1, else 0**

This intends to find all the agregated cells (4x4) where both land covers are found (10 from min and 30 from max). This will give us the areas where we can make comparison

```{r}
# # create mask: if A + B = 40 and D - C > 100 then 1 else 0
# mask <- ((A + B == 40) & (C - D < 200)) * 1

# create mask: if A + B = 40 and D - C > 100 then 1 else 0
mask <- ((A + B == 40)) * 1

plot(mask)

#writeRaster(mask, "mask_forest_low_veg_pairs.tif", overwrite = TRUE)
```

Convert the mask pixels (pixels containing pairs) to polygons with unique ids

```{r}
#Convert each pixel to a polygon
mask1 <- mask
mask1[mask1 != 1] <- NA   # all other pixels become NA

polygons <- as.polygons(mask1, dissolve = FALSE)
# Optionally, assign an ID to each polygon
polygons$ID <- 1:nrow(polygons)
# Inspect
polygons
```

Save them locally (optional)

```{r}
# Save as GeoJSON
writeVector(polygons, "mask_forest_low_veg_pair.geojson", filetype = "GeoJSON", overwrite = TRUE)
```

## 3) Extract paired points

To do this we will first convert all the landcover pixels which fall in the masked raster into points

```{r}
#first, convert the masked raster to the same resolution as the original landcover raster
mask <- resample(mask, ldc_filtered, method = "near")  # nearest neighbor

# then apply the mask to the landcover data. This will leave only the areas where we have pairs.
ldc_mask <- ldc_filtered*mask
plot(ldc_mask, col = c("white","forestgreen","goldenrod"))

# turn all zeros to NULL
ldc_mask[ldc_mask == 0] <- NA

#convert all remaining pixels to points
ldc_pts <- as.points(ldc_mask, na.rm = TRUE)

#write the id of each polygon to the points
ldc_pts$poly_id <- extract(polygons, ldc_pts)[, "ID"]

#Attach landcover class from the raster
ldc_pts$landcover <- extract(ldc_mask, ldc_pts)[,2]  # 2nd column = raster values

print(ldc_pts)


```

Add attributes of elevation and UID

```{r}

ldc_pts$elv <- extract(elv, ldc_pts)[,2]
ldc_pts$uid <- seq_len(nrow(ldc_pts)) 
ldc_pts
nrow(ldc_pts)
```

Now we will filter only those elevation categories for which we intend to do our LST analyses

```{r}

# Extract attribute table as data frame
df <- as.data.frame(ldc_pts)

# Leave only those points at the chosen elevation category
df_filtered <- df %>%
  filter( 
    (elv >= 0 & elv <= 100) |
    (elv >= 600 & elv <= 700) |
    (elv >= 1300 & elv <= 1500)
  )
  
# Now subset the original vector to keep only those rows
pts_filtered <- ldc_pts[df_filtered$uid, ]   # or use ID column if present
pts_filtered
```

Now we will remove the duplicate points within one polygon.

```{r}
# Get only the attribute table (does not touch geometry)
att <- values(pts_filtered)

# Find duplicates by poly_id + landcover
keep <- !duplicated(att[c("poly_id", "landcover")])

# Subset the SpatVector
pts_unique <- pts_filtered[keep, ]
pts_unique

```

And remove any single points, where there is not a pair anymore

```{r}
# Get only the attribute table (does not touch geometry)
att_u <- values(pts_unique)

# Count how many points per poly_id
poly_counts <- table(att_u$poly_id)

# Keep only poly_ids with more than one point
keep_ids <- names(poly_counts[poly_counts > 1]) 

# Subset the SpatVector
pts_unique <- pts_unique[att_u$poly_id %in% keep_ids, ]

pts_unique
```

## 4) Get 10 pairs of spatially balanced points for each elevation

With the code above we have obtained:

-   218 pairs at low elevation (0-100m)

-   28 pairs at medium elevation (600-700m)

-   20 pairs at high elevation (1300-1500)

Now we want to extract 10 pairs of spatially balanced samples for each elevation class

**4.1) Divide the data in our three categories**

We will extract the polygons we defined in step 2, which contain our pairs. Then we will select 10 polygons based on the spatially balanced method (grts) for each elev category. After that we will select only those points from points_unique that have the same poly_id as our spatially balanced polygons.

```{r}

# Extract attributes
att <- values(pts_unique)

# Devide the points data in their elevation category. 
pts_low  <- pts_unique[att$elv <= 100, ]
pts_mid  <- pts_unique[att$elv >= 600 & att$elv <= 700, ]
pts_high <- pts_unique[att$elv >= 1300, ]

print(nrow(pts_low)/2)
print(nrow(pts_mid)/2)
print(nrow(pts_high)/2)

# Function to extract polygons matching point poly_ids
get_polys_for_points <- function(pts, polys, pts_id = "poly_id", poly_id = "ID") {
  # Extract unique IDs from points
  ids <- unique(values(pts)[[pts_id]])
  # Subset polygons using their ID field
  polys_subset <- polys[values(polys)[[poly_id]] %in% ids, ]
  return(polys_subset)
}

# Get the polygons which have those points
# In addition we take the centers as points and convert them from spat vector to sf object. This extra steps make it easier when working with the grts function in spsurvey later
polys_low_sf <- st_as_sf(centroids(get_polys_for_points(pts_low, polygons)))
polys_mid_sf <- st_as_sf(centroids(get_polys_for_points(pts_mid, polygons)))
polys_high_sf <- st_as_sf(centroids(get_polys_for_points(pts_high, polygons)))

print(nrow(polys_low_sf))
print(nrow(polys_mid_sf))
print(nrow(polys_high_sf))


```

**4.2) Get spatially balanced samples**

```{r}
library(spsurvey)
set.seed(51)
```

```{r}
# Now we will extract the spatially balanced data with the grts function from spsurvey for each elevation class

# # TEST
# poly_high_grts <- (grts(polys_high_sf, n_base = 10))
# plot(poly_high_grts, polys_high_sf, key.width = lcm(3))

# We add $sites_base to convert the result in sf object
# Apply to all the point datasets
poly_low_grts <- (grts(polys_low_sf, n_base = 10))$sites_base
poly_mid_grts <- (grts(polys_mid_sf, n_base = 10))$sites_base
poly_high_grts <- (grts(polys_high_sf, n_base = 10))$sites_base



```

Extract the ids of the spatially balanced polygons

```{r}
ids_low_grts <- poly_low_grts$ID
ids_mid_grts <- poly_mid_grts$ID
ids_high_grts <- poly_high_grts$ID

# print(length(unique(ids_low_grts)))
# print(length(unique(ids_mid_grts)))
# print(length(unique(ids_high_grts)))
```

```{r}
pts_low_grts <- pts_low[values(pts_low)$poly_id %in%ids_low_grts, ]
pts_mid_grts <- pts_mid[values(pts_mid)$poly_id %in%ids_mid_grts, ]
pts_high_grts <- pts_high[values(pts_high)$poly_id %in%ids_high_grts, ]

```

## 5) Export the data

```{r}

# # all pairs generated
# writeVector(pts_unique, "DATA/pairs_timseries_v260123_1432.shp",
#             filetype = "ESRI Shapefile", overwrite = TRUE)

# spatially balanced 10 pairs for each elevation category
writeVector(pts_low_grts, "DATA/pairs_timseries_low_10.shp",
            filetype = "ESRI Shapefile", overwrite = TRUE)

writeVector(pts_mid_grts, "DATA/pairs_timseries_mid_10.shp",
            filetype = "ESRI Shapefile", overwrite = TRUE)

writeVector(pts_high_grts, "DATA/pairs_timseries_high_10.shp",
            filetype = "ESRI Shapefile", overwrite = TRUE)
```
