---
title: "00_preprocess_raw_data_layers"
format: html
---

# Prepossessing of raw input data

```{r}
# for spatial layers and era5 data
library(terra)
library(ggplot2)
library(dplyr)
library(here)
library(stringr)
library(readr)

# for LST data cleaning
library(lubridate)
library(tidyverse)
library(terra)
library(dplyr)
library(scales)

source(here("01_functions", "functions.R"))
```

## Spatial Layers projections and extent

Reproject the data to the same projection. For Costa Rica, we have chosen EPSG:3261761

```{r}

# this is our area of interest, it has been provided in already processed form, (projection 32617) done in ArcGIS software
aoi <- vect(here("00_data", "processed", "Amistosa_climate_hub_32617.shp"))
# ----------------------------------------
# load landcover data (categorical raster)
# https://doi.org/10.5281/zenodo.13329296
ldc <- rast(here("00_data", "raw", "8_Amostosa_landcover_YN.tif"))

plot(ldc, type="classes", col = rainbow(nrow(ldc)))

# reproject the data
ldc <- project(ldc, "EPSG:32617", method = "near")

# save the processed data
writeRaster(
  ldc, here("00_data", "processed", "landcover_10_2023_32617.tif"),overwrite=TRUE)
  
# -----------------------------------------
# load elevation data (continuous raster)
# NASA SRTM Digital Elevation 30m downloaded for the region of interest from GEE

elv <- rast(here("00_data", "raw", "elevation_30_YN.tif" ))

elv <- project(elv, "EPSG:32617")
writeRaster(
  elv, here("00_data","processed","elevation_30_32617.tif"),overwrite=TRUE)

# ---------------------------------------
# maximum water extent 1984–2024
# https://doi:10.1038/nature20584 

mwe <- rast(here("00_data","raw", "extent_90W_10N_v1_5_2024.tif")) 

mwe <- project(mwe, "EPSG:32617")

# We will crop the mwe  according to our area of interest, to save memory, since it is too big

mwe_crop <- crop(x = mwe, y = aoi)

writeRaster(
  mwe_crop, here("00_data","processed","water_extent_30_32617.tif"), overwrite=TRUE)

# -------------------------------------
# Load open street map rivers
rivers <- vect(here("00_data", "raw", "hotosm_cri_waterways_lines_shp.shp")) 
rivers <- project(rivers, "EPSG:32617")

# crop to aoi extent
rivers_crop <- crop(x = rivers, y = aoi)


# save the projected vector
writeVector(
  rivers_crop, here("00_data", "processed", "rivers_osm_32617.shp")
  )
# -------------------------------------
#  Canopy height
# the canopy height has been downloaded in tiles from Planet explorer and mosaiced in QGIS with EPSG:32617 projection
```

## ERA5 2m air temperature

```{r}
# country boundaries
base <- vect (here("00_data", "vis_layers", "world-administrative-boundaries.geojson")) |>
  project("EPSG:32617") |>
  crop(ext(aoi))

plot(base)

plot(base[1,])
plot(base[2,]) #background
plot(base[3,])

# aggregate costa rica and panama land borders, base[2, ] is a background poly
b_union <- terra::union(base[1,],base[3,]) 
b_union <- terra::aggregate(b_union, fun = "mean", dissolve = TRUE)

plot(b_union) 

ocean <- erase(base[2,], b_union)
plot(ocean)

# -----------------------------------------------------------
# load the grib file from ERA5 monthly data
era5 <- rast(here("00_data", "99_temp", "era5_2m_temp_1900UTC_1330solar.grib")) |>
  project("EPSG:32617", method = "near")

# crate a mask layer that masks everything outside of the aoi, and does not allow any pixels touching ocean

r <- era5[[1]] #raster of same res as era5,
values(r) <- 1 # but with only 1 cell values

mask1 <- mask(r, ocean, touches=TRUE, updatevalue=0) 
mask1 <- 1 - mask1 #invert the values
plot(mask1)

mask2 <- mask(r, aoi, touches=FALSE, updatevalue=0)
plot(mask2)

mask <- mask1*mask2
mask[mask == 0] <- NA

plot (mask)
plot(aoi, add = TRUE)

# --------------------------
# #write a tif file for visualization later
# writeRaster(
#   mask,
#   here("00_data", "vis_layers", "aoi_mask_era5.tif"),
#   overwrite = TRUE
# )
# ---------------------------
era5_masked <- era5*mask
plot(era5_masked[[1]])
plot(aoi, add = TRUE)
```

```{r}


# load the grib file from ERA5 monthly data
meangrib <- function(path, mask) {
  r <- rast(path)
  # project
  r <- project(r, "EPSG:32617", method = "near") 
  # remove pixels outside the aoi, using the precreated mask
  r <- r*mask
  
}
# load the grib file from ERA5 monthly data
# readgrib <- function(path, mask) {
t1330_r <- meangrib(here("00_data", "99_temp", "era5_2m_temp_1900UTC_1330solar.grib"), mask)
plot(t1330_r[[1]])
# ------------------
# #write a tif file for visualization later
# writeRaster(
#   t1330_r[[1]],
#   here("00_data", "vis_layers", "ERA5_pixel_extent.tif"),
#   overwrite = TRUE
# )
#-------------------------

t0130_r <- meangrib(here("00_data", "99_temp", "era5_2m_temp_0700UTC_0130solar.grib"), mask)
plot(t0130_r[[1]])

t1330_mean <- global(t1330_r, "mean", na.rm = TRUE)

t0130_mean <- global(t0130_r, "mean", na.rm = TRUE)
# ------------------------------------

# join in a tibble
climate <- tibble(
  # date = as.Date(time(temp)),
  t1330 = t1330_mean[, 1],
  t0130 = t0130_mean[, 1],
  band_t1330 = names(t1330_r),
  band_t0130 = names(t0130_r)
  # n_month = as.numeric(str_extract(names(t1300), "\\d+$"))
  ) |> 
  # arrange(n_month) |>
  mutate(
    date = seq(
      as.Date("2020-01-01"),
      by = "month",
      length.out = n()
    )
  )

climate

# -------------------------------
climate <- climate |>
    filter(
    date >= as.Date("2020-10-01"), 
    date < as.Date("2024-11-01")
  ) |>
  mutate(
    t1330 = t1330 - 273.15,
    t0130 = t0130 - 273.15
    # precipitation = precipitation * 1000
  ) |>
  select(c("date", "t1330", "t0130"))

climate

# Save the monthly table for fast reuse
write_csv(climate, here("00_data","processed", "t2m_1330_0130_solar_time.csv"), append = FALSE)
```

## ERA5 precipitation data

-\> we have downloaded the hourly data and are manually merging it together, because of an official issue reported on the Copernicus portal

```{r}

r <- rast(here("00_data", "99_temp", "era5_total_precipitation_hourly.nc"))
r <- project(r, "EPSG:32617", method = "near") 
  # remove pixels outside the aoi, using the precreated mask
r <- r*mask

tp_hourly <- global(r, fun = "mean", na.rm = TRUE)
tp_hourly

tp_hourly$date_time <- time(r)
tp_hourly
# --------------------------------
library(dplyr)
library(lubridate)

monthly_tp <- tp_hourly |>
  mutate(
    date = floor_date(date_time, "month")
  ) |>
  group_by(date) |>
  summarise(
    precipitation = sum(mean, na.rm = TRUE)
  )
monthly_tp
# --------------------------------

monthly_tp <- monthly_tp |>
  mutate(
    precipitation_mm = precipitation * 1000,
  ) |>
    filter(
    date >= as.Date("2020-10-01"), 
    date < as.Date("2024-11-01")
  )|>
  select(c("date", "precipitation_mm"))

# Save the monthly table for fast reuse
write_csv(monthly_tp, here("00_data","processed", "monthly_total_precip.csv"), append = FALSE)
```

## ERA5 solar incoming

Surface solar radiation downwards (J m^-2^ ) - Amount of solar radiation (also known as shortwave radiation) reaching the surface of the Earth.

```{r}
mask <- rast(here("00_data", "vis_layers", "aoi_mask_era5.tif"))
plot(mask)
r <- rast(here("00_data", "99_temp", "era5_solar_radiation_1900UTC_1330solar.nc"))
r <- project(r, "EPSG:32617", method = "near") 
  # remove pixels outside the aoi, using the precreated mask
r <- r*mask

plot(r[[1]])
names(r)
time(r)

ssrd_1330_mean <- global(r, "mean", na.rm = TRUE)

ssrd_1330_mean
# ------------------------------------

# join in a tibble
ssrd <- tibble(
  # date = as.Date(time(temp)),
  ssrd_Jm2_1330 = (ssrd_1330_mean[, 1]), #devided by 3600 to convert Jm2 to Wm2 for one hour of observation
  band_ssrd_1330 = names(r),
  # n_month = as.numeric(str_extract(names(t1300), "\\d+$"))
  ) |> 
  # arrange(n_month) |>
  mutate(
    date = seq(
      as.Date("2020-01-01"),
      by = "month",
      length.out = n()
    ),
    ssrd_Wm2_1330 = ssrd_Jm2_1330/3600 #devided by 3600 to convert Jm2 to Wm2 for one hour of observation
  )

ssrd

# -------------------------------
ssrd <- ssrd |>
    filter(
    date >= as.Date("2020-10-01"), 
    date < as.Date("2024-11-01")
  ) |>
  select(c("date", "ssrd_Jm2_1330", "ssrd_Wm2_1330"))

ssrd
#!!!! Note ssrd values seem waj too big for an hour. Check later, nevertheless the relationship is linear, so the pattern should still show on

# Save the monthly table for fast reuse
write_csv(ssrd, here("00_data","processed", "ssrd_1330_solar_time.csv"), append = FALSE)
```

## LST data

```{r}
## LST data
  
df01 <- clean_lst_data(here("00_data", "raw", "pairs_timeseries_results_0130.csv"))
df01

df13 <- clean_lst_data(here("00_data", "raw", "pairs_timeseries_results_1330.csv"))
df13

```

```{r}

# Calculate the cooling for each pair (with identical poly_id) throughout the study period. Cooling is expressed as the LST difference between open land and forest.
df13_cool <- cooling(df13)
df13_cool

df01_cool <- cooling(df01)
df01_cool

# # Save the monthly table for fast reuse
# write_csv(df13_cool, here("00_data", "processed", "lst_cooling_pairs_daily_1330.csv"), append = FALSE)
# 
# write_csv(df01_cool, here("00_data", "processed", "lst_cooling_pairs_daily_0130.csv"), append = FALSE)

```

```{r}

# check if elevation is the same for all poly_ids (should be the same)
check_elv <- function(df) {
  df|>
    mutate(
      year  = year(date),
      month = month(date)
    ) |>
    group_by(poly_id, year, month) |>
    summarise(n_elv = n_distinct(elv_mean)) |>
    filter(n_elv > 1)
}

check_elv(df13_cool)
check_elv(df01_cool)
```

```{r}
#Monthly calculations

# calculate the monthly cooling for each pair
cooling_pair_monthly <- function (df_cool) {
  df_cool |>
    mutate(
      year  = year(date),
      month = month(date)
    ) |>
    group_by(poly_id, year, month) |>
    summarise(
      forest = mean(forest, na.rm = TRUE),
      open_land = mean(open_land, na.rm = TRUE),
      lst_mean = mean(lst_mean, na.rm = TRUE),
      cooling = mean(cooling, na.rm = TRUE),
      elv_mean = unique(elv_mean), #get the elv value, there should be a unique value for each poly id. If not an error will come back
      n_points = n()    # optional (how many records per month)
    ) |>
    ungroup() |>
    # optional: create a proper first-of-month date
    mutate(month_date = as.Date(sprintf("%04d-%02d-01", year, month)))
}

monthly13_pairs <- cooling_pair_monthly(df13_cool)
monthly13_pairs
monthly01_pairs <- cooling_pair_monthly(df01_cool)
monthly01_pairs

# # Save the monthly table for fast reuse
# write_csv(monthly13_pairs, here("00_data", "processed", "lst_cooling_pairs_monthly_1330.csv"), append = FALSE)
# 
# write_csv(monthly01_pairs, here("00_data", "processed", "lst_cooling_pairs_monthly_0130.csv"), append = FALSE)


monthly01_pairs <- read_csv(here("00_data", "processed", "lst_cooling_pairs_monthly_0130.csv"))

```

```{r}
#calculate the monthly mean accross all pairs
cooling_mean_monthly <- function (monthly_pairs) {
   monthly_pairs |>
    group_by(year, month) |>
    summarise(
      forest = mean(forest, na.rm = TRUE),
      open_land = mean(open_land, na.rm = TRUE),
      lst_mean = mean(lst_mean, na.rm = TRUE),
      cooling = mean(cooling, na.rm = TRUE),
      n_points = n()    # optional (how many records per month)
    ) |>
    ungroup() |>
  mutate(month_date = as.Date(sprintf("%04d-%02d-01", year, month)))
}

monthly01_mean <- cooling_mean_monthly(monthly01_pairs)
monthly01_mean
monthly13_mean <- cooling_mean_monthly(monthly13_pairs)
monthly13_mean

# Save the monthly table for fast reuse
write_csv(monthly01_mean, here("00_data", "processed", "lst_cooling_monthly_mean_0130.csv"))

write_csv(monthly13_mean, here("00_data", "processed", "lst_cooling_monthly_mean_1330.csv"))

```

```{r}

# print the hottest months for open land according to the lst data
hottest_months <- monthly13_mean |>
  group_by(year) |>
  slice_max(lst_mean, n = 1, with_ties = FALSE) |>
  ungroup()

hottest_months

#1330
# year  month forest open land lst mean cooling   n_points
# 2021	3	28.77358	33.87539	31.32448	5.101806	64	
# 2022	2	29.63851	34.58259	32.11055	4.944076	64
# 2023	2	30.94451	36.23678	33.59065	5.292268	64	
# 2024	3	31.63097	38.71252	35.17174	7.081551	64	

#0130
# year  month forest open land lst mean cooling   n_points
# 2021	4	15.20531	19.47772	17.34152	4.272408	64
# 2022	3	15.24137	20.11842	17.67990	4.877056	64
# 2023	4	16.16014	21.06052	18.61033	4.900375	64
# 2024	3	17.25689	22.24100	19.74895	4.984114	64

```
