---
title: "cooling"
format: html
editor_options: 
  chunk_output_type: inline
---

# Load prerequisites

```{r}
library(dplyr)
library(tidyr)
library(readr)
library(here)
library(ggstatsplot)
library(ggplot2)
library(hrbrthemes)

monthly13_pairs <- read_csv(here("00_data", "processed", "lst_cooling_pairs_monthly_1330.csv"))

monthly01_pairs <- read_csv(here("00_data", "processed", "lst_cooling_pairs_monthly_0130.csv"))


# filter to include only the hottest months (based on monthly air temp)
hottest_month_tair_01 <- monthly01_pairs |>
  filter(
    (year == 2021 & month == 3) |
    (year == 2022 & month == 2) |
    (year == 2023 & month == 2) |
    (year == 2024 & month == 3)
  ) |>
  mutate(year = factor(year))

hottest_month_tair_13 <- monthly13_pairs |>
  filter(
    (year == 2021 & month == 3) |
    (year == 2022 & month == 2) |
    (year == 2023 & month == 2) |
    (year == 2024 & month == 3)
  ) |>
  mutate(year = factor(year))

```

# Plot Cooling ggbetweenstats

```{r}

# plot dif for hottest month Tair
ggbetweenstats(
  data = hottest_month_tair_13,
  x = year, 
  y=cooling,
  bf.message = FALSE,
  results.subtitle = FALSE
) 

ggbetweenstats(
  data = hottest_month_tair_01,
  x = year, 
  y=cooling,
  bf.message = FALSE,
  results.subtitle = FALSE
) 

# compare forest and open land
ggbetweenstats(
  data = hottest_month_tair_13,
  x = year, 
  y=forest,
  bf.message = FALSE,
  results.subtitle = FALSE
) 

ggbetweenstats(
  data = hottest_month_tair_13,
  x = year, 
  y=open_land,
  bf.message = FALSE,
  results.subtitle = FALSE
) 
```

# Statistics

### Prepare the data for the statistical tests

#### Cooling

```{r}
# First: reshape your data into paired format for 2021 and 2024 comparison
df01_wide <- hottest_month_tair_01 |>
  select(poly_id, year, cooling,elv_mean) |>
  pivot_wider(names_from = year, values_from = cooling)

df01_wide


#--------------------------------------
df13_wide <- hottest_month_tair_13 |>
  select(poly_id, year, cooling,elv_mean) |>
  pivot_wider(names_from = year, values_from = cooling)

df13_wide

# write_csv(df_wide, "df_wide.csv")

```

```{r}
# Quick sanity check
df13_wide_test <- df13_wide |>
  mutate(diff = `2024` - `2021`)

summary(df13_wide_test$diff)
hist(df13_wide_test$diff)

# !!!! difference distribution seems slightly skewed. Check what to do in this case
```

#### Forest

```{r}
# First: reshape your data into paired format for 2021 and 2024 comparison
df01_wide_forest <- hottest_month_tair_01 |>
  select(poly_id, year, forest,elv_mean) |>
  pivot_wider(names_from = year, values_from = forest)

df01_wide_forest


#--------------------------------------
df13_wide_forest <- hottest_month_tair_13 |>
  select(poly_id, year, forest,elv_mean) |>
  pivot_wider(names_from = year, values_from = forest)

df13_wide_forest

```

#### Open Land

```{r}
df13_wide_open_land <- hottest_month_tair_13 |>
  select(poly_id, year, open_land,elv_mean) |>
  pivot_wider(names_from = year, values_from = open_land)

df13_wide_open_land

```

### Hypothesis (important)

Because you *expect 2024 \> 2021*, this is a **one-sided paired test**.

**Null hypothesis (H₀):**

H0: μ2024−μ2021=0

(no change in cooling )

**Alternative hypothesis (H₁):**

H1: μ2024−μ2021\>0

(2024 is warmer on average)

#### Compare El Nino year (2024) to all reference years (2021, 2022, 2023)

```{r}

df = df13_wide_open_land

# library(tidyverse)
# Run the paired t-test in R
test1 <- t.test(df$`2024`, df$`2021`,
       paired = TRUE,
       alternative = "greater")

# Run the paired t-test in R
test2 <- t.test(df$`2024`, df$`2022`,
       paired = TRUE,
       alternative = "greater")

# Run the paired t-test in R
test3 <- t.test(df$`2024`, df$`2023`,
       paired = TRUE,
       alternative = "greater")
test1
test2
test3
# Bind the p values all together to perform multiple testing
p <- c(test1$p.value, test2$p.value, test3$p.value)
test1$p.value
test2$p.value
test3$p.value

# perform multiple testing with Bonferony Holm
p.adjust(p, method = "holm", n = length(p))
```

#### Compare all reference years (2021, 2022, 2023) among each other

```{r}
# test and plot the difference between the reference years
df_cool_reference_years <- hottest_month_tair_13 |>
  filter(
    (year == 2021) |
    (year == 2022) |
    (year == 2023)
#     (year == 2024 & month == 4)
  )

# Welch's one-way ANOVA
oneway.test(
  cooling ~ year,
  data =df_cool_reference_years,
  var.equal = FALSE
)



```

# Plot elevations

```{r}

plot_cooling_vs_elevation <- function(df, method = "lm") {

  # Determine smoothing method and formula
  if (method == "poly") {
    smooth_method <- "lm"
    smooth_formula <- y ~ poly(x, 2)
    plot_title <- "Polynomial model"
  } else {
    smooth_method <- method
    smooth_formula <- NULL
    plot_title <- switch(
      method,
      lm = "Linear model",
      loess = "Loess model",
      gam = "GAM model",
      paste(toupper(method), "model")
    )
  }

  p <- ggplot(df, aes(x = elv_mean, y = cooling)) +
    geom_point() +
    geom_smooth(
      method = smooth_method,
      formula = smooth_formula,
      color = "red",
      fill = "#69b3a2",
      se = TRUE
    ) +
  scale_y_continuous(
    limits = c(-2, 14),
    breaks = seq(-2, 14, by = 2)
  ) +
    labs(
      x = "Elevation (m)",
      y = "Forest Cooling (°C)",
      title = plot_title
    ) +
    facet_wrap(~year, ncol = 4) +
    theme_ipsum()

  return(p)
}


# Linear
plot_cooling_vs_elevation(hottest_month_tair_13, "lm")

# Polynomial (quadratic)
plot_cooling_vs_elevation(hottest_month_tair_13, "poly")

# Loess
plot_cooling_vs_elevation(hottest_month_tair_13, "loess")

# GAM (if desired)
plot_cooling_vs_elevation(hottest_month_tair_13, "gam")

#----------------------------------------
# Linear
# plot_cooling_vs_elevation(hottest_month_tair_01, "lm")
# 
# # Polynomial (quadratic)
# plot_cooling_vs_elevation(hottest_month_tair_01, "poly")
# 
# # Loess
# plot_cooling_vs_elevation(hottest_month_tair_01, "loess")
# 
# # GAM (if desired)
# plot_cooling_vs_elevation(hottest_month_tair_01, "gam")
```
