---
title: "pairs_timeseries"
format: html
editor_options: 
  chunk_output_type: console
---

```{r}
library(lubridate)
library(tidyverse)
library(terra)
library(dplyr)
library(scales)
library(here)

source(here("01_functions", "functions.R"))
```

# Raw LST data cleaning and visualization

## Calculation

```{r}
# # Read into a tiblle
# csv <- read_csv(here("04_output", "pairs_timeseries_results.csv"))
# 
# # Clean the raw data
# df <- clean_lst_data(csv)
# df

```

```{r}


```

## Plot the data

```{r}
# Plot the timeseries
p <- plot_raw (df, "0-1800")
p

# Save the plots
# Map(ggsave,
  #   filename = c("media/p1.png", "media/p2.png", "media/p3.png"),
  #   plot = list(p1, p2, p3),
  #   width = 250,
  #   height = 150,
  #   units = "mm"
  # )
```

```{r}
# plot each separate class in a facet
ggplot(df, aes(x = date, y = lst, color = landcover)
  ) +
    # geom_point(alpha = 0.4) +            # optional: show the raw points
    geom_jitter(size = 0.3, alpha = 0.25) +
    geom_smooth(method = "loess",        # "loess" for local smoothing
                span = 0.1,              # adjust smoothness (0-1)
                se = FALSE,
                fill = "#4e4f52",
                linewidth = 0.9) +               # display confidence interval
    scale_colour_manual(
      values = c("darkgreen", "sienna2")
    ) +
    scale_y_continuous(limits = c(5, 35), breaks = seq(5, 35, by = 5)
    ) +
    # scale_x_date(date_breaks = "2 months",  date_labels = "%Y %b")+
    # scale_y_continuous(limits = c(5, 40), breaks = seq(5, 40, by = 5))+
    annotate("rect", xmin = as.Date("2023-11-01"), xmax = as.Date("2024-04-01"), ymin = -Inf, ymax = Inf, alpha = .2
    ) +
    annotate("text", x = as.Date("2023-12-01"), y = 30, label = "El Niño"
    ) +
    labs(
      x = "Date", 
      y = "LST (°C)", 
      title = paste("Daily LST time-series for forest and open land at","0-1800", "m")
    ) +
    facet_wrap(~landcover) +
    theme_classic()
```

### 

```{r}
# Facet grids for each separate polygon
plot_facet<- function (df, elev_range) {
  ggplot(df, aes(x = date, y = lst, color = landcover)
  ) +
    # geom_point(alpha = 0.4) +            # optional: show the raw points
    geom_jitter(size = 0.5, alpha = 0.25) +
    geom_smooth(method = "loess",        # "loess" for local smoothing
                span = 0.1,              # adjust smoothness (0-1)
                se = FALSE,
                fill = "#4e4f52",
                linewidth = 0.6) +               # display confidence interval
    scale_colour_manual(
      values = c("darkgreen", "sienna2")
    ) +
    scale_y_continuous(limits = c(5, 35), breaks = seq(5, 35, by = 5)
    ) +
    guides(x = guide_axis(angle = 45)) +
    # scale_x_date(date_breaks = "2 months",  date_labels = "%Y %b")+
    # scale_y_continuous(limits = c(5, 40), breaks = seq(5, 40, by = 5))+
    annotate("rect", xmin = as.Date("2023-11-01"), xmax = as.Date("2024-04-01"), ymin = -Inf, ymax = Inf,         alpha = .2
    ) +
    # annotate("text", x = as.Date("2023-12-01"), y = 30, label = "El Niño"
    # ) +
    labs(
      x = "Date", 
      y = "LST (°C)", 
      title = paste("Daily LST time-series for forest and open land at",elev_range, "m")
    ) + facet_wrap(~poly_id, ncol = 5) +
    theme_minimal()
    
}



# facet <- plot_facet(df, "0-1800")
# ggsave(
#   "facet.png",
#     facet,
#     width = 250,
#     height = 900,
#     units = "mm"
# )
```

# Monthly LST

## Calculation

`{# {r} # # apply function monthly_lst() to get the monthly value for each point # monthly <- monthly_lst(df) # monthly #  # # Save the monthly table for fast reuse # write_csv(monthly, here("04_output", "monthly_lst.csv")) #  # # print the hottest months for open land according to the lst data # hottest_months <- monthly |> #   filter(landcover == "open_land") |> #   group_by(year) |> #   slice_max(lst_monthly_mean, n = 1, with_ties = FALSE) |> #   ungroup() #  # hottest_months # # # results # # landcover  year month lst_monthly_mean n_points month_date # # open_land  2021     4             19.5     1216 2021-04-01 # # open_land  2022     3             20.1     1280 2022-03-01 # # open_land  2023     4             21.0     1226 2023-04-01 # # open_land  2024     4             22.3     1216 2024-04-01 #`

## Plot monthly

```{r}
# Plot monthly values
p_month <- plot_monthly(monthly,"0-1800m")
p_month
```

```{r}
# plot with confidence interval
library(dplyr)
library(lubridate)

monthly_lst <- function(df) {
  df |>
    mutate(
      year  = year(date),
      month = month(date)
    ) |>
    group_by(landcover, year, month) |>
    summarise(
      mean_lst = mean(lst, na.rm = TRUE),
      sd_lst   = sd(lst, na.rm = TRUE),
      n_points = sum(!is.na(lst)),
      .groups = "drop"
    ) |>
    mutate(
      se = sd_lst / sqrt(n_points),
      ci_lower = mean_lst - 1.96 * se,
      ci_upper = mean_lst + 1.96 * se,
      month_date = as.Date(sprintf("%04d-%02d-01", year, month))
    )
}

monthly <- monthly_lst(df)

library(ggplot2)

plot_monthly <- function(df, elev_range) {

  ggplot(
    df,
    aes(
      x = month_date,
      y = mean_lst,
      color = landcover,
      fill = landcover
    )
  ) +

    geom_ribbon(
      aes(
        ymin = ci_lower,
        ymax = ci_upper
      ),
      alpha = 0.1,
      linewidth = 0.0,
      show.legend = FALSE
    ) +

    geom_line(linewidth = 0.6) +

    scale_colour_manual(
      values = c("darkgreen", "sienna2")
    ) +

    scale_fill_manual(
      values = c("darkgreen", "sienna2")
    ) +

    scale_x_date(
      date_breaks = "3 months",
      date_labels = "%b %Y"
    ) +

    scale_y_continuous(
      limits = c(10, 30),
      breaks = seq(10, 30, by = 5)
    ) +

    annotate(
      "rect",
      xmin = as.Date("2023-11-01"),
      xmax = as.Date("2024-04-01"),
      ymin = -Inf,
      ymax = Inf,
      alpha = .2
    ) +

    annotate(
      "text",
      x = as.Date("2023-12-01"),
      y = 28,
      label = "El Niño"
    ) +

    labs(
      x = "Month",
      y = "LST",
      title = paste(
        "Monthly mean LST time-series for forest and open land at",
        elev_range, "m"
      )
    ) +

    theme_classic()+
    theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
}

p_month <- plot_monthly(monthly,"0-1800m")
p_month
```

# Change analyses with the bfastmonitor() function

Looking at Oct 2020 - Oct 2022 as a reference (no el Nino conditions) and detecting whether any breaks occur after that

```{r}
library(bfast)

df_monthly <- monthly

# Choose one landcover - forest or open land
open_land_monthly_summarize <- df_monthly |>  filter(landcover == "open_land")
open_land_monthly_summarize

forest_monthly_summarize <- df_monthly |>  filter(landcover == "forest")
forest_monthly_summarize
# ------------------------------------------------------------------
# stl --> seasonal trend decomposition with loess
ts_o <- ts(open_land_monthly_summarize$lst_monthly_mean, frequency = 12, start = c(2020,10))
plot(stl(ts_o,"per"))

ts_f <- ts(forest_monthly_summarize$lst_monthly_mean, frequency = 12, start = c(2020,10))
plot(stl(ts_f,"per"))

# ------------------------------------------------------------------
# bfast

plot(bfast(ts_o, season = "harmonic", breaks = 1))
plot(bfast(ts_f, season = "harmonic", breaks = 1))

# ------------------------------------------------------------------
# bfastmonitor

bfm <- function(df){
  ts <- ts(df$lst_monthly_mean, frequency = 12, start = c(2020,10))
  bfm <- bfastmonitor(ts, formula = response ~ harmon, start = c(2022,10))
}

plot(bfm(open_land_monthly_summarize),ylab = "LST", ylim = c(14,28), sub = "Open Land at 0-1800m")
(bfm(open_land_monthly_summarize))$magnitude

plot(bfm(forest_monthly_summarize),ylab = "LST", ylim = c(14,28), sub = "Forest at 0-1800m")
(bfm(forest_monthly_summarize))$magnitude
# # |>
# #   select(c(lst))
#   
# ts <- ts(open_land_monthly_summarize$lst, frequency = 12, start = c(2020,10))
# ts
# 
# 
# bfm <- bfastmonitor(ts, formula = response ~ harmon, start = c(2022,10))
# plot(bfm,ylab = "LST", ylim = c(16,28), sub = "Open Land at 0-100m")
# 
# ts(1:10, frequency = 4, start = c(1959,1), end = 1962) # 2nd Quarter of 1959
```

```{r}
# 
# source("functions.R")
# 
# Map(plot_bfast_change,
#     list(df_monthly, df_monthly),
#     c("forest", "open_land"),
#     c("0-1800m", "0-1800m")
#     )

```

# Monthly Anomaly

## Calculation Mean across all points

```{r}
# df <- low_df

plot_monthly_anom <- function (df, df_monthly, elev_range) {

  #--------------------------------------
  # Get the monthly mean for the first two reference years for each pair points
  monthly_ref_means <- df |>
  filter(date >= as.Date("2020-10-01"),
         date <= as.Date("2022-10-31")
  ) |>
  # extract month number
  mutate(month = month(date)
  ) |>
  # compute the mean for each month (across all years in the window)
  group_by(landcover, month, poly_id) |>
  summarise(
    ref_mean = mean(lst, na.rm = TRUE),
    .groups = "drop"
  )
  
  #--------------------------------------
  #Get the monthly mean for each pair accross the whole period
  df_monthly <- df |>
    mutate(
      year  = year(date),
      month = month(date)
    ) |>
    group_by(landcover,year, month, poly_id) |>
    summarise(
      lst_monthly_mean = mean(lst, na.rm = TRUE),
      n_points = n(),    # optional (how many records per month)
    ) |>
    ungroup() |>
    # optional: create a proper first-of-month date
    mutate(month_date = as.Date(sprintf("%04d-%02d-01", year, month)))

  #--------------------------------------
  df_monthly_anom <- df_monthly |>
  left_join(monthly_ref_means, by = c("landcover", "month", "poly_id")) |>
  mutate(anomaly = lst_monthly_mean - ref_mean )
  
#--------------------------------------
  plot_anom<- function (df, elev_range) {
  ggplot(
    df,
    aes(
      x = month_date,
      y = anomaly,
      color = landcover)
    )+
    # geom_line(
    #   aes(color = landcover),
    #   linewidth = 0.6
    # ) +
    geom_jitter(size = 0.6, alpha = 0.5) +
    geom_smooth(method = "loess",# "loess" for local smoothing
                span = 0.1,      # adjust smoothness (0-1)
                se = FALSE,
                fill = "#4e4f52",
                linewidth = 1) +
    scale_colour_manual(
      values = c("darkgreen", "sienna2")
    ) +
    scale_colour_manual(
      values = c("darkgreen", "sienna2")
    ) +
    scale_y_continuous(
      limits = c(-3, 8), 
      breaks = seq(-2, 8, by = 2)
      )+
    annotate("rect", xmin = as.Date("2023-11-01"), xmax = as.Date("2024-04-01"), ymin = -Inf, ymax = Inf, alpha = .2
    ) +
    annotate("text", x = as.Date("2023-12-01"), y = 28, label = "El Niño"
    ) +
    geom_hline(
      yintercept = 0, 
      color = "#4e4f52", 
      linetype = "dashed", 
      linewidth = 0.8) +
      annotate("rect", xmin = as.Date("2023-11-01"), xmax = as.Date("2024-04-01"), ymin = -Inf, ymax = Inf, alpha = .2
    )+
    labs(
      x = "Date",
      y = "LST anomaly (°C)",
      title = paste("Monthly LST anomaly of forest and open land at",elev_range, "m")
    )+
    theme_minimal()
}

plot_anom(df_monthly_anom,elev_range)

}


plot_monthly_anom (df, df_monthly, "0-1800m")
```

## Facets: Pair wise

```{r}
# df <- low_df

plot_monthly_anom <- function (df, df_monthly, elev_range) {

  #--------------------------------------
  # Get the monthly mean for the first two reference years for each pair points
  monthly_ref_means <- df |>
  filter(date >= as.Date("2020-10-01"),
         date <= as.Date("2022-10-31")
  ) |>
  # extract month number
  mutate(month = month(date)
  ) |>
  # compute the mean for each month (across all years in the window)
  group_by(landcover, month, poly_id) |>
  summarise(
    ref_mean = mean(lst, na.rm = TRUE),
    .groups = "drop"
  )
  
  #--------------------------------------
  #Get the monthly mean for each pair accross the whole period
  df_monthly <- df |>
    mutate(
      year  = year(date),
      month = month(date)
    ) |>
    group_by(landcover,year, month, poly_id) |>
    summarise(
      lst_monthly_mean = mean(lst, na.rm = TRUE),
      n_points = n(),    # optional (how many records per month)
    ) |>
    ungroup() |>
    # optional: create a proper first-of-month date
    mutate(month_date = as.Date(sprintf("%04d-%02d-01", year, month)))

  #--------------------------------------
  df_monthly_anom <- df_monthly |>
  left_join(monthly_ref_means, by = c("landcover", "month", "poly_id")) |>
  mutate(anomaly = lst_monthly_mean - ref_mean )
  
#--------------------------------------
  plot_anom<- function (df, elev_range) {
  ggplot(
    df,
    aes(
      x = month_date,
      y = anomaly,
      color = landcover)
    )+
    geom_line(
      aes(color = landcover),
      linewidth = 0.6
    ) +
    geom_jitter(size = 0.6, alpha = 0.25) +
    # geom_smooth(method = "loess",        # "loess" for local smoothing
    #             span = 0.5,              # adjust smoothness (0-1)
    #             se = FALSE,
    #             fill = "#4e4f52",
    #             linewidth = 0.9) +               # display confidence interval
    scale_colour_manual(
      values = c("darkgreen", "sienna2")
    ) +
    scale_colour_manual(
      values = c("darkgreen", "sienna2")
    ) +
    scale_y_continuous(
      limits = c(-3, 8), 
      breaks = seq(-2, 8, by = 2)
      )+
    annotate("rect", xmin = as.Date("2023-11-01"), xmax = as.Date("2024-04-01"), ymin = -Inf, ymax = Inf, alpha = .2
    ) +
    annotate("text", x = as.Date("2023-12-01"), y = 28, label = "El Niño"
    ) +
    geom_hline(
      yintercept = 0, 
      color = "#4e4f52", 
      linetype = "dashed", 
      linewidth = 0.8) +
      annotate("rect", xmin = as.Date("2023-11-01"), xmax = as.Date("2024-04-01"), ymin = -Inf, ymax = Inf, alpha = .2
    )+
    labs(
      x = "Date",
      y = "LST anomaly (°C)",
      title = paste("Monthly LST anomaly of forest and open land at",elev_range, "m")
    ) + facet_wrap(~poly_id, ncol = 5) +
    theme_minimal()
}

plot_anom(df_monthly_anom,elev_range)

}


# Map(plot_monthly_anom,
#     list(low_df, mid_df, high_df),
#     list(low_monthly, mid_monthly, high_monthly),
#     c("0-100m","600-700", "1300-1500")
# )

facet_anom <- plot_monthly_anom (df, df_monthly, "0-1800m")


# ggsave(
#   "facet_anom.png",
#     facet_anom,
#     width = 250,
#     height = 900,
#     units = "mm"
# )
```

# Cooling

## Calculation

### Full timeseries

```{r}
library(lubridate)
library(tidyverse)
library(terra)
library(dplyr)
library(scales)
library(here)
source(here("01_functions", "functions.R"))

library(readr)
# # Read into a tiblle
# csv <- read_csv(here("04_output", "pairs_timeseries_results.csv"))
# 
# # Clean the raw data
# df <- clean_lst_data(csv)
# df
# 
# # Calculate the cooling for each pair (with identical poly_id) throughout the study period. Cooling is expressed as the LST difference between open land and forest.
# df_cool <- df |>
#   pivot_wider(
#     id_cols = c(poly_id, date),
#     names_from = landcover,
#     values_from = lst
#   ) |>
#   mutate(
#     forest_cooling = open_land - forest #NOTE! cooling is usealy expressed in negative terms as fores - open land. Here we calculate it in positive terms as this helps in the following analyses and stat tests 
#   )
# 
# df_cool

```

### Monthly cooling

```{r}
# # calculate the monthly cooling
# df_cool_month <- df_cool |>
#     mutate(
#       year  = year(date),
#       month = month(date)
#     ) |>
#     group_by(poly_id, year, month) |>
#     summarise(
#       cooling_monthly_mean = mean(forest_cooling, na.rm = TRUE),
#       n_points = n(),    # optional (how many records per month)
#     ) |>
#     ungroup() |>
#     # optional: create a proper first-of-month date
#     mutate(month_date = as.Date(sprintf("%04d-%02d-01", year, month))
#            )
```

### Add elevation

Compare cooling accross elevations. Add elevation for each poly_id based on the mean elevation value from forest and open land

```{r}
# 
# # load the points data and convert it to a dataframe
# df_pts <- vect(here("00_data", "data_selection", "pairs_32617.shp")) |>
#   as.data.frame()
# 
# # set the landcover as a factor, as now it is just a code (10 and 30)
# df_pts$landcover <- factor(
#   df_pts$landcover,
#   levels = c(10, 30),
#   labels = c("forest", "open_land")
# )
# 
# df_pts
# 
# # calculate the mean elevation for each poly_id
# poly_elv <- df_pts |>
#   pivot_wider(
#     id_cols = poly_id,
#     names_from = landcover,
#     values_from = elv
#    )|>
#   mutate(
#     poly_elv = (open_land + forest)/2
#   )
# 
# poly_elv
# 
# # Join the table containing the elevation for each polygon with the table containing the cooling values
# df_cool_month <- df_cool_month |>
#   left_join(
#     poly_elv |> select(poly_id, poly_elv),
#     by = "poly_id"
#   )
# 
# df_cool_month
```

```{r}
# Save the cooling table for fast reuse
# write_csv(df_cool_month, here("04_output", "df_cool_month.csv"))
```

# Cooling v2

## Find hottest date

```{r}
# # apply function monthly_lst() to get the monthly value for each point
# monthly <- monthly_lst(df)
# monthly
# 
# # Save the monthly table for fast reuse
# write_csv(monthly, here("04_output", "monthly_lst.csv"))
# 
# # print the hottest months for open land according to the lst data
# hottest_months <- monthly |>
#   filter(landcover == "open_land") |>
#   group_by(year) |>
#   slice_max(lst_monthly_mean, n = 1, with_ties = FALSE) |>
#   ungroup()
# 
# hottest_months
# # # results
# # landcover  year month lst_monthly_mean n_points month_date
# # open_land  2021     4             19.5     1216 2021-04-01
# # open_land  2022     3             20.1     1280 2022-03-01
# # open_land  2023     4             21.0     1226 2023-04-01
# # open_land  2024     4             22.3     1216 2024-04-01
# 
```

```{r}
print("hello")
```

```{r}

# df_cool_month <- read_csv(here("00_data", "processed", "lst_cooling_pairs_monthly.csv"))
# 
# 
# 
# # filter to include only the hottest months (based on monthly lst for open land)
# df_cool_hot_month <- df_cool_month |>
#   filter(
#     (year == 2021 & month == 3) |
#     (year == 2022 & month == 2) |
#     (year == 2023 & month == 2) |
#     (year == 2024 & month == 3)
#   )
# 
# df_cool_hot_month$year <- factor(df_cool_hot_month$year)
# 
# df_cool_hot_month
```

### Plot the data using boxplots

```{r}
# Libraries

# library(hrbrthemes)
# library(viridis)
# 
# 
# df_cool_hot_month |>
#   ggplot( aes(x = factor(year), y=cooling_monthly_mean, fill=factor(year))) +
#     geom_boxplot() +
#     scale_fill_viridis_d( alpha=0.6) +
#   scale_y_continuous(breaks = seq(floor(min(df_cool_month$cooling_monthly_mean, na.rm = TRUE)),
#                                 ceiling(max(df_cool_month$cooling_monthly_mean, na.rm = TRUE)),
#                                 by = 2)) +
#     geom_jitter(color="black", size=0.4, alpha=0.9) +
#     theme_ipsum() +
#     theme(
#       legend.position="none",
#       plot.title = element_text(size=11)
#     ) +
#   xlab("")+
#   ylab("Forest cooling (°C)")
```

```{r}
library(dplyr)
library(readr)
library(here)
library(ggstatsplot)

monthly13_pairs <- read_csv(here("00_data", "processed", "lst_cooling_pairs_monthly_1330.csv"))

monthly01_pairs <- read_csv(here("00_data", "processed", "lst_cooling_pairs_monthly_0130.csv"))

# filter to include only the hottest months (based on monthly air temp)
hottest_month_tair_01 <- monthly01_pairs |>
  filter(
    (year == 2021 & month == 3) |
    (year == 2022 & month == 2) |
    (year == 2023 & month == 2) |
    (year == 2024 & month == 3)
  ) |>
  mutate(year = factor(year))

hottest_month_tair_13 <- monthly13_pairs |>
  filter(
    (year == 2021 & month == 3) |
    (year == 2022 & month == 2) |
    (year == 2023 & month == 2) |
    (year == 2024 & month == 3)
  ) |>
  mutate(year = factor(year))



# plot dif for hottest month Tair
ggbetweenstats(
  data = hottest_month_tair_13,
  x = year, 
  y=cooling,
  bf.message = FALSE,
  results.subtitle = FALSE
)
```

```{r}
# For comparison with precipitation and solar as well
# library(ggstatsplot)
# 
# df_cool_month <- read_csv(here("00_data", "processed", "lst_cooling_pairs_monthly.csv"))
# 
# 
# 
# # filter to include only the hottest months (based on monthly air temp)
# hottest_month_tair <- df_cool_month |>
#   filter(
#     (year == 2021 & month == 3) |
#     (year == 2022 & month == 2) |
#     (year == 2023 & month == 2) |
#     (year == 2024 & month == 3)
#   ) |>
#   mutate(year = factor(year))
# 
# hottest_month_tair
# #--------------------------------------------------------------
# # filter to include only the hottest months (based on monthly solar radiation)
# hottest_month_rad <- df_cool_month |>
#   filter(
#     (year == 2021 & month == 3) |
#     (year == 2022 & month == 3) |
#     (year == 2023 & month == 4) |
#     (year == 2024 & month == 3)
#   ) |>
#   mutate(year = factor(year))#--------------------------------------------------------------
# # filter to include only the hottest months (based on min precipitation)
# hottest_month_precip <- df_cool_month |>
#   filter(
#     (year == 2021 & month == 3) |
#     (year == 2022 & month == 1) |
#     (year == 2023 & month == 2) |
#     (year == 2024 & month == 1)
#   ) |>
#   mutate(year = factor(year))
# 
# 
# 
# # plot dif for hottest month Tair
# ggbetweenstats(
#   data = hottest_month_tair,
#   x = year, 
#   y=cooling,
#   bf.message = FALSE,
#   results.subtitle = FALSE,
#   
# )
# #--------------------------------------------------------------
# # plot dif for hottest month rad
# ggbetweenstats(
#   data = hottest_month_rad,
#   x = year, 
#   y=cooling,
#   bf.message = FALSE,
#   results.subtitle = FALSE,
#   
# )
# 
# # plot dif for hottest month precip
# ggbetweenstats(
#   data = hottest_month_precip,
#   x = year, 
#   y=cooling,
#   bf.message = FALSE,
#   results.subtitle = FALSE,
#   
# )
```

### Plot cooling time-series: Mean across all pairs

\`\`\`{# {r} \# \# plot_forest_cooling \<- function(df, elev_range) { \#\
\# df\<-df_cool \# \# \# ggplot(df_cool, aes(x = date, y = forest_cooling) \# )+ \# geom_jitter(size = 0.5, alpha = 0.25 \# )+ \# geom_smooth(method = "loess", \# "loess" for local smoothing \# span = 0.1, \# adjust smoothness (0-1) \# se = FALSE, \# fill = "#4e4f52", \# linewidth = 0.8, \# color = 'black' \# )+ \# geom_smooth(method=lm, color = 'darkred' \# )+ \# geom_hline( \# yintercept = 0, \# color = "#4e4f52", \# linetype = "dashed", \# linewidth = 0.8 \# )+ \# scale_y_continuous( \# limits = c(-2.5, 10), \# breaks = seq(-2.5, 10, by = 2.5) \# )+ \# annotate("rect", xmin = as.Date("2023-11-01"), xmax = as.Date("2024-04-01"), ymin = -Inf, ymax = Inf, alpha = .2 \# )+ \# labs( \# x = "Date", \# y = "Forest Cooling (°C)", \# title = paste("Forest induced cooling at",elev_range, "m") \# )+ \# theme_minimal() \# } \# \# plot_forest_cooling(df,"0-1800m")

```         

### Plot cooling: For each pair

```{# {r}
# 
# plot_forest_cooling_pairs <- function(df, elev_range){ 
#   df<-df_cool
# 
#   ggplot(df_cool, aes(x = date, y = forest_cooling)
#     )+
#     geom_jitter(size = 0.5, alpha = 0.25
#     )+
#     geom_smooth(method = "loess",        # "loess" for local smoothing
#                   span = 0.1,              # adjust smoothness (0-1)
#                   se = FALSE,
#                   fill = "#4e4f52",
#                   linewidth = 0.8,
#                 color = 'black'
#     )+
#     geom_smooth(method=lm, color = 'darkred'
#     )+
#     geom_hline(
#       yintercept = 0, 
#       color = "#4e4f52", 
#       linetype = "dashed", 
#       linewidth = 0.8) +
#       annotate("rect", xmin = as.Date("2023-11-01"), xmax = as.Date("2024-04-01"), ymin = -Inf, ymax = Inf, alpha = .2
#     )+
#     scale_y_continuous(
#       limits = c(-2.5, 10), 
#       breaks = seq(-2.5, 10, by = 2.5)
#     )+
#     labs(
#       x = "Date",
#       y = "Forest Cooling (°C)",
#       title = paste("Forest induced cooling at",elev_range, "m")
#     )+
#     facet_wrap(~poly_id, ncol = 5
#     )+
#     theme_bw()
#   }
# 
# facet_cool <- plot_forest_cooling_pairs(df,"0-1800m")
# 
# ggsave(
#   "facet_cool.png",
#     facet_cool,
#     width = 250,
#     height = 900,
#     units = "mm"
# )
```

### Plot the data - elevation

```{r}
p4 <- ggplot(hottest_month_tair, aes(x=elv_mean, y=cooling)) +

  geom_point() +

  geom_smooth(method=lm ,formula = y ~ poly(x, 2), color="red", fill="#69b3a2", se=TRUE) +

  scale_y_continuous(breaks = seq(floor(min(hottest_month_tair$cooling, na.rm = TRUE)),

                                ceiling(max(hottest_month_tair$cooling, na.rm = TRUE)),

                                by = 2))+

  facet_wrap(~year, ncol = 4) +

  theme_ipsum() 

p4
```

```{r}
# linear trend + confidence interval
p3 <- ggplot(hottest_month_tair, aes(x=elv_mean, y=cooling)) +
  geom_point() +
  geom_smooth(method=lm , color="red", fill="#69b3a2", se=TRUE) +
  scale_y_continuous(breaks = seq(floor(min(hottest_month_tair$cooling, na.rm = TRUE)),
                                ceiling(max(hottest_month_tair$cooling, na.rm = TRUE)),
                                by = 2))+
    labs(
      x = "Elevation (m)",
      y = "Forest Cooling (°C)",
      title = "Linear model") +
  facet_wrap(~year, ncol = 4) +
  theme_ipsum() 

p3
# Non linear - polynomial
p4 <- ggplot(hottest_month_tair, aes(x=elv_mean, y=cooling)) +
  geom_point() +
  geom_smooth(method=lm ,formula = y ~ poly(x, 2), color="red", fill="#69b3a2", se=TRUE) +
  scale_y_continuous(breaks = seq(floor(min(hottest_month_tair$cooling, na.rm = TRUE)),
                                ceiling(max(hottest_month_tair$cooling, na.rm = TRUE)),
                                by = 2))+
  labs(
      x = "Elevation (m)",
      y = "Forest Cooling (°C)",
      title = "Polynomial model") +
  facet_wrap(~year, ncol = 4) +
  theme_ipsum() 

p4

# Non linear - loess
p4 <- ggplot(hottest_month_tair, aes(x=elv_mean, y=cooling)) +
  geom_point() +
  geom_smooth(method=loess, color="red", fill="#69b3a2", se=TRUE) +
  scale_y_continuous(breaks = seq(floor(min(hottest_month_tair$cooling, na.rm = TRUE)),
                                ceiling(max(hottest_month_tair$cooling, na.rm = TRUE)),
                                by = 2))+
    labs(
      x = "Elevation (m)",
      y = "Forest Cooling (°C)",
      title = "Loess model") +
  facet_wrap(~year, ncol = 4) +
  theme_ipsum() 

p4
```

## Statistical test

### Prepare the data for the statistical tests

```{r}
# First: reshape your data into paired format for 2021 and 2024 comparison
df01_wide <- hottest_month_tair_01 |>
  filter(year %in% c(2021, 2022, 2023, 2024)) |>
  select(poly_id, year, cooling,elv_mean) |>
  pivot_wider(names_from = year, values_from = cooling)

df01_wide

#--------------------------------------
df13_wide <- hottest_month_tair_13 |>
  filter(year %in% c(2021, 2022, 2023, 2024)) |>
  select(poly_id, year, cooling,elv_mean) |>
  pivot_wider(names_from = year, values_from = cooling)

df13_wide

# write_csv(df_wide, "df_wide.csv")

# Quick sanity check
df01_wide_test <- df_wide |>
  mutate(diff = `2024` - `2021`)

summary(df01_wide_test$diff)
hist(df01_wide_test$diff)

# !!!! difference distribution seems slightly skewed. Check what to do in this case

```

### Hypothesis (important)

Because you *expect 2024 \> 2021*, this is a **one-sided paired test**.

**Null hypothesis (H₀):**

H0: μ2024−μ2021=0

(no change in LST)

**Alternative hypothesis (H₁):**

H1: μ2024−μ2021\>0

(2024 is warmer on average)

### Compare all reference years (2021, 2022, 2023) to El Nino year (2024)

Results Paired t-test ( alternative hypothesis: true mean difference is greater than 0)

df_wide\$\`2024\` and df_wide\$\`2021\` t = 6.1645, df = 63, **p-value = 2.781e-08** 95 percent confidence interval: 0.7999085 Inf \| sample estimates: mean difference **1.096978**

data: df_wide\$\`2024\` and df_wide\$`2022` t = 8.3698, df = 63, **p-value = 4.03e-12** alternative hypothesis: true mean difference is greater than 0 95 percent confidence interval: 1.199686 Inf \| sample estimates: mean difference **1.498586**

data: df_wide\$\`2024\` and df_wide\$`2023` t = 6.721, df = 63, **p-value = 3.054e-09** alternative hypothesis: true mean difference is greater than 0 95 percent confidence interval: 0.9396064 Inf sample estimates: mean difference **1.250118**

**Adjusted p-values with Holm:**

```         
2.781018e-08 1.209015e-11 6.108294e-09
```

```{r}
library(tidyverse)
# Run the paired t-test in R
test1 <- t.test(df_wide01$`2024`, df_wide01$`2021`,
       paired = TRUE,
       alternative = "greater")

# Run the paired t-test in R
test2 <- t.test(df_wide01$`2024`, df_wide01$`2022`,
       paired = TRUE,
       alternative = "greater")

# Run the paired t-test in R
test3 <- t.test(df_wide01$`2024`, df_wide01$`2023`,
       paired = TRUE,
       alternative = "greater")
test1
test2
test3
# Bind the p values all together to perform multiple testing
p <- c(test1$p.value, test2$p.value, test3$p.value)

# perform multiple testing with Bonferony Holm
p.adjust(p, method = "holm", n = length(p))
```

### Compare all reference years (2021, 2022, 2023) among each other

```{r}

# test and plot the difference between the reference years
df_cool_reference_years <- hottest_month_tair |>
  filter(
    (year == 2021) |
    (year == 2022) |
    (year == 2023)
#     (year == 2024 & month == 4)
  )

# Welch's one-way ANOVA
oneway.test(
  cooling ~ year,
  data =df_cool_reference_years,
  var.equal = FALSE
)

# # -------------------------------
# # Test the reference years also using pairwise comparisons
# # Run the paired t-test in R
# test4 <- t.test(df_wide$`2021`, df_wide$`2022`,
#        paired = TRUE,
#        alternative = "two.sided")
# 
# # Run the paired t-test in R
# test5 <- t.test(df_wide$`2022`, df_wide$`2023`,
#        paired = TRUE,
#        alternative = "two.sided")
# 
# # Run the paired t-test in R
# test6 <- t.test(df_wide$`2021`, df_wide$`2023`,
#        paired = TRUE,
#        alternative = "two.sided")
# 
# # Bind the p values all together to perform multiple testing
# p_ref <- c(test4$p.value, test5$p.value, test6$p.value)
# 
# # perform multiple testing with Bonferony Holm
# p.adjust(p_ref, method = "holm", n = length(p_ref))

```
